---
title: 'Shallow Water, Smaller Waves: How Depth Limits Australia''s Biggest Storm
  Waves'
author: "Julian O'Grady"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    highlight: tango
    fig_caption: true
subtitle: "A simple CCHaPS model of depth, latitude and 1-year and 40-year extreme
  significant wave heights"
params:
  datadir: "C:/Users/ogr013/OneDrive - CSIRO/Documents/ACS_NCRA/schism/"
  minimum_depth_m: 0.5
  maximum_depth_m: 80
  reference_latitude: -25
  map_resolution_degrees: 0.0125
  map_gap_fill_passes: 3
  wave_display_max_m: 18
  residual_display_max_m: 5
  residual_display_increment_m: 0.5
  cores: 8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  fig.align = "center",
  out.width = "100%",
  dpi = 150
)

required_packages <- c(
  "terra", "ncdf4", "loopevd", "parallel", "leaflet", "htmltools", "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install the following packages before knitting: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(terra)
library(ncdf4)
library(loopevd)
library(leaflet)

available_cores <- parallel::detectCores(logical = FALSE)
if (!is.finite(available_cores)) available_cores <- 1
analysis_cores <- max(1, min(params$cores, available_cores))

colz <- rev(c(
  "#000078", "#0000BB", "#0000FF", "#0C3AFC", "#1975FA", "#25B0F7",
  "#32EBF5", "#73EEE6", "#B4F1D7", "#F5F5C8", "#F5F564", "#F5E032",
  "#F6CC00", "#F7A300", "#F97A00", "#FB5100", "#FF0000", "#DA0006",
  "#B5000D", "#910014", "#7B003D", "#660066", "#88008C", "#AB00B2",
  "#CD00D8", "#F000FF", "#CAA6DE", "#DBA7DD", "#EDA9D2"
))

colzRamp <- grDevices::colorRampPalette(rev(colz))
bwr <- grDevices::colorRampPalette(c(
  "#000078", "blue", "blue", "white", "white", "red", "red", "#7B003D"
))

wave_breaks <- seq(0, params$wave_display_max_m, by = 1)
residual_breaks <- seq(
  -params$residual_display_max_m,
  params$residual_display_max_m,
  by = params$residual_display_increment_m
)

input_files <- list(
  bathymetry = file.path(
    params$datadir,
    "depth_cchaps_hindcast_BARRA-R2_WHACS_ERA5_fx.nc"
  ),
  mean_water_level = file.path(
    params$datadir,
    "timeseries",
    "zos_cchaps_hindcast_BARRA-R2_WHACS_ERA5_yr_mean_198101010000-202012310000.nc"
  ),
  hs_gumbel = file.path(
    params$datadir,
    "EVA",
    "hs_cchaps_hindcast_BARRA-R2_WHACS_ERA5_yr_max_198101010000-202012310000_fgumbel.nc"
  ),
  coastline = file.path(
    params$datadir,
    "LGA_2023_AUST_GDA2020_coastline"
  )
)
```

<style>
.plain-language {
  background: #e9f7fc;
  border-left: 7px solid #28a9d6;
  border-radius: 5px;
  color: #18394a;
  font-size: 1.17em;
  line-height: 1.58;
  margin: 1.4em 0;
  padding: 1.05em 1.25em 1.05em 3.8em;
  position: relative;
}
.plain-language::before {
  align-items: center;
  background: #178fc0;
  border-radius: 50%;
  color: white;
  content: "i";
  display: flex;
  font-family: Georgia, serif;
  font-size: 1.25em;
  font-weight: bold;
  height: 1.75em;
  justify-content: center;
  left: 1.05em;
  position: absolute;
  top: 1.05em;
  width: 1.75em;
}
.plain-language strong:first-child {
  display: block;
  font-size: 1.06em;
  margin-bottom: 0.2em;
}
.lightbox-overlay {
  align-items: center;
  background: rgba(0, 0, 0, 0.88);
  display: none;
  inset: 0;
  justify-content: center;
  padding: 3.5rem 2rem 2rem;
  position: fixed;
  z-index: 99999;
}
.lightbox-overlay.open { display: flex; }
.lightbox-overlay img {
  background: white;
  cursor: default;
  max-height: 91vh;
  max-width: 96vw;
  object-fit: contain;
}
.lightbox-close {
  background: transparent;
  border: 0;
  color: white;
  cursor: pointer;
  font-size: 3rem;
  line-height: 1;
  position: absolute;
  right: 1.1rem;
  top: 0.45rem;
}
.main-container img:not(.leaflet-tile) { cursor: zoom-in; }
.leaflet-container img { cursor: inherit !important; }
.leaflet-control-layers,
.leaflet-control-layers label,
.leaflet-control.legend {
  font-size: 14px;
  line-height: 1.28;
}
.caption-note {
  color: #4f5c65;
  font-size: 0.94em;
}
</style>

<div id="figure-lightbox" class="lightbox-overlay" role="dialog" aria-modal="true" aria-label="Enlarged figure">
  <button class="lightbox-close" type="button" aria-label="Close enlarged figure">&times;</button>
  <img src="" alt="Enlarged report figure">
</div>

# From the global wave climate to Australia's shallow shelves {.unnumbered}

This regional analysis follows the companion explainer [`Global_Hs_AEP_latitude_model_comparison.Rmd`](https://ausclimateservice.github.io/latitude-and-big-waves/Global_Hs_AEP_latitude_model_comparison.html), which asks how much of the global pattern of extreme significant wave height can be represented by latitude alone. That global analysis found that a smooth latitude-only relationship represented more than half of the mapped variation for both the 1-year and 40-year ARIs. Latitude therefore captures an important first-order control: the broad organisation of wind belts, storm tracks and tropical versus mid-latitude wave climates.

The global maps also showed substantial differences between places at the same latitude. Ocean-basin width, storm exposure, swell pathways, land shelter and regional extreme-generating systems create organised residual patterns that latitude cannot represent. Those differences were generally larger for the 40-year ARI. Near coasts, some negative residuals were also consistent with shallow continental shelves limiting the larger incoming waves.

This report takes that final observation as its starting point. It moves from a global, smooth latitude relationship to the much finer Australian CCHaPS mesh and asks how local water depth changes the regional pattern. The latitude term used below is deliberately linear over Australia, while depth is added explicitly as a predictor.

# The question

Why are the largest storm waves smaller over shallow continental shelves, gulfs and banks than in nearby deeper water? This report uses the Australian Climate Service **Coupled Coastal Hazard Prediction System (CCHaPS)** to examine how strongly local water depth is associated with extreme significant wave height, $H_s$.

The analysis compares two event sizes:

- a **1-year average recurrence interval (ARI)** wave height, representing a relatively common annual extreme; and
- a **40-year ARI** wave height, representing a rarer and generally larger event.

It then replaces the full numerical wave model with a deliberately simple model that knows only local water depth and latitude. Depth represents the shallow-water constraint, while latitude provides a first-order guide to Australia's broad north–south storm climate. Comparing the simple model with CCHaPS shows what these two controls can explain and what they miss.

Much of the remaining structured variance reflects physics that is represented by numerical wave modelling but absent from this two-predictor relationship. CCHaPS accounts for processes such as wind input, swell propagation, shoaling, refraction, sheltering by coasts and reefs, wave–current interaction, bottom friction, whitecapping and depth-induced breaking. Regional storm tracks and tropical cyclones also change the wave field arriving at each shelf. The residuals therefore show where the full process-based model departs from what water depth and latitude alone would predict.

## A quick guide to the jargon of an ocean-wave climate scientist

| Term | Plain-language meaning |
|:--|:--|
| Significant wave height, $H_s$ | The average height of the highest one-third of waves. It describes a sea state, not the single largest wave. |
| Water depth, $h$ | The average vertical distance between the seabed and water surface at a CCHaPS model node. |
| ARI | Average recurrence interval. A 40-year ARI does **not** mean a storm lasts 40 years; it describes how rare the estimated wave height is. |
| AEP | Annual exceedance probability: the chance that a level is exceeded in any year. |
| Depth-limited breaking | Wave-energy loss that occurs when waves become too large for the available water depth. |
| Residual | CCHaPS wave height minus the depth-and-latitude estimate. Positive values mean CCHaPS is higher than the simple model predicts. |
| $R^2$ | The fraction of variation represented by the fitted relationship. It measures fit, not proof of causation. |

**Who might find this useful?** This is an educational touchstone for surfers, students, educators and readers curious about the fundamental physical controls on coastal wave climate. It is not a local surf forecast, navigation product or engineering-design assessment.

# How can depth limit a wave?

In deep water, waves can travel without interacting strongly with the seabed. As waves move into shallower water, the seabed increasingly changes their speed, direction and shape. Shoaling can initially increase wave height, but bottom friction and depth-induced breaking eventually remove energy and prevent waves from growing without limit.

A common physical shorthand is that the limiting height of an individual wave is some fraction of the local depth:

\[
H_{\max} \approx \gamma h,
\]

where $h$ is water depth and γ is a breaker index. Here, $H_{\max}$ refers to an upper limit for an **individual wave**, not a maximum value of significant wave height, $H_s$. Significant wave height describes the distribution of waves in a sea state—approximately the average height of the highest one-third—so it cannot be substituted directly for $H_{\max}$.

This distinction is particularly important during extreme storm winds. In the Battjes–Janssen formulation, random wave heights are represented by a truncated distribution and only a fraction of the waves, $Q_b$, are breaking at a given location. The bulk loss of wave energy depends on that breaking fraction as well as the limiting individual-wave height. A storm sea can therefore have a large $H_s$ without every wave reaching $H_{\max}$. The value of γ is also not universal: it depends on wave period, seabed slope, wave spectrum and the wave-height definition. Consequently, the $H_s/h$ ratios examined below are useful depth-limitation diagnostics but are not direct estimates of γ.

CCHaPS does not impose the simple equation above as its final answer. Its WWMIII wave component solves the spectral wave-action balance and includes wind input, wave interactions, whitecapping, bottom friction and depth-limited breaking. Depth-limited breaking is represented using the established [Battjes and Janssen (1978) approach](https://doi.org/10.1061/9780872621909.034).

<div class="plain-language">
<strong>In plain language</strong>
A very large individual wave cannot remain that large when there is too little water beneath it: it steepens, breaks and loses energy. In a storm, however, only part of the full distribution of waves may be breaking at once. Significant wave height summarises the whole sea state, so its depth limit is not identical to the limit for the single largest wave. Depth is also not the whole story—two places with the same depth can experience very different waves because their storm exposure, fetch, sheltering and incoming swell are different.
</div>

# The CCHaPS model and data

CCHaPS is a national coupled circulation-and-wave model built on an unstructured coastal mesh. The hindcast covers 1981–2020 and simulates tides, surge, currents and waves around Australia, with finer resolution near the coast. Its wave component is forced by downscaled atmospheric reanalysis and spectral waves entering the model boundary.

The hindcast, model description and extreme-value outputs are available from:

- [CCHaPS Hindcast for Australia](https://data.csiro.au/collection/csiro%3A65669);
- [Extreme Value Analysis of CCHaPS](https://data.csiro.au/collection/csiro%3A65867); and
- [Hernaman et al. (2026), CCHaPS development, implementation and validation](https://doi.org/10.1016/j.ocemod.2026.102731).

This report uses modelled $H_s$, bathymetric depth and mean modelled sea-surface height at each node. Average water depth is calculated as:

\[
h = h_{\mathrm{bathymetry}} + \overline{\mathrm{zos}},
\]

where bathymetry is positive downward and the overbar denotes the 1981–2020 mean sea-surface height relative to the model datum.

# Estimating 1-year and 40-year storm waves

The attached exploratory script calculated several pre-selected AEP levels. Here the same fitted Gumbel distributions are used to calculate the exact AEP corresponding to each requested ARI:

\[
\mathrm{AEP}(T) = 1 - \exp\left(-\frac{1}{T}\right).
\]

This gives an AEP of 63.2% for a 1-year ARI and 2.47% for a 40-year ARI. The 40-year value describes a rare annual maximum; it is not a prediction that the event will occur on a fixed 40-year schedule.

The report uses three deliberately simple models. A latitude-only benchmark is fitted across **all valid positive water depths**:

\[
H_s = \alpha_0 + \alpha_\phi(\phi-\phi_0).
\]

The response remains in metres for the latitude relationship: latitude is not fitted to $\log_{10}(H_s)$. The depth-only model restores the power-law formulation from the exploratory analysis and is fitted over the shelf-depth range from `r params$minimum_depth_m` to `r params$maximum_depth_m` m:

\[
\log_{10}(H_s) = \log_{10}(C) + \beta_d\log_{10}(h),
\]

which is equivalent to $H_s=C h^{\beta_d}$. To keep latitude linear on the original wave-height scale, the combined model uses that fitted depth exponent as a fixed power-law shape and estimates an additive latitude effect in metres:

\[
H_s = \beta_0 + C h^{\beta_d} + \beta_\phi(\phi-\phi_0).
\]

Here \(\phi\) is latitude and \(\phi_0\) is a reference latitude of `r abs(params$reference_latitude)`°S. The exponent \(\beta_d\) is estimated from the log–log depth fit; then \(\beta_0\), \(C\) and \(\beta_\phi\) are estimated by an ordinary linear model on $H_s$ in metres. Thus, the depth response retains the original power-law shape while the latitude correction remains additive and linear rather than becoming exponential. The all-depth latitude-only fit is a broader benchmark, so its performance values should not be compared as if it used the same fitting domain.

Latitude is not treated as a physical cause by itself. It is a deliberately simple proxy for large-scale changes in storm type and wave exposure around Australia; regional fetch, coastline orientation, sheltering and tropical-cyclone exposure remain outside the model.

```{r load-calculate-and-fit, results='hide', fig.keep='none'}
# Check the local inputs before performing the expensive calculations.
missing_inputs <- names(input_files)[!vapply(input_files, file.exists, logical(1))]
if (length(missing_inputs) > 0) {
  stop(
    "These inputs were not found beneath params$datadir: ",
    paste(missing_inputs, collapse = ", "),
    ". Update the YAML datadir parameter or the input filenames."
  )
}

# Read the static CCHaPS mesh and bathymetry.
bathy_nc <- ncdf4::nc_open(input_files$bathymetry)
bathy_longitude <- ncdf4::ncvar_get(bathy_nc, "SCHISM_hgrid_node_x")
bathy_latitude <- ncdf4::ncvar_get(bathy_nc, "SCHISM_hgrid_node_y")
bathymetry_m <- ncdf4::ncvar_get(bathy_nc, "depth")
ncdf4::nc_close(bathy_nc)

# Add the 1981-2020 mean water-level offset to bathymetric depth.
zos_nc <- ncdf4::nc_open(input_files$mean_water_level)
zos_yearly_mean <- ncdf4::ncvar_get(zos_nc, "zos_yr_mean")
ncdf4::nc_close(zos_nc)

node_means <- function(x, number_of_nodes) {
  dimensions <- dim(x)
  if (is.null(dimensions)) {
    if (length(x) != number_of_nodes) stop("Unexpected zos array length.")
    return(as.numeric(x))
  }
  if (dimensions[1] == number_of_nodes) return(rowMeans(x, na.rm = TRUE))
  if (length(dimensions) >= 2 && dimensions[2] == number_of_nodes) {
    return(colMeans(x, na.rm = TRUE))
  }
  stop("The zos array cannot be aligned with the CCHaPS bathymetry nodes.")
}

mean_zos_m <- node_means(zos_yearly_mean, length(bathymetry_m))
mean_zos_m[!is.finite(mean_zos_m)] <- NA_real_
average_depth_m <- bathymetry_m + mean_zos_m

# Read the fitted Gumbel parameters used by the original depth analysis.
evd_with_coordinates <- loopevd::netcdf_to_df(
  input_files$hs_gumbel,
  exclude_cov = FALSE
)

longitude_name <- intersect(
  names(evd_with_coordinates),
  c("lon", "longitude", "SCHISM_hgrid_node_x")
)[1]
latitude_name <- intersect(
  names(evd_with_coordinates),
  c("lat", "latitude", "SCHISM_hgrid_node_y")
)[1]

if (is.na(longitude_name) || is.na(latitude_name)) {
  stop("Longitude and latitude columns were not recognised in the EVA file.")
}

longitude <- evd_with_coordinates[[longitude_name]]
latitude <- evd_with_coordinates[[latitude_name]]
evd_parameters <- evd_with_coordinates[
  setdiff(names(evd_with_coordinates), c(longitude_name, latitude_name))
]

if (length(average_depth_m) != nrow(evd_parameters)) {
  stop("Bathymetry and EVA files contain different numbers of CCHaPS nodes.")
}

coordinate_difference <- max(
  abs(longitude - bathy_longitude),
  abs(latitude - bathy_latitude),
  na.rm = TRUE
)
if (!is.finite(coordinate_difference) || coordinate_difference > 1e-7) {
  stop("The bathymetry and EVA files do not use the same node order.")
}

# Remove invalid fitted distributions before calculating return levels.
valid_evd <- rep(TRUE, nrow(evd_parameters))
if ("loc" %in% names(evd_parameters)) {
  valid_evd <- valid_evd & is.finite(evd_parameters$loc) & evd_parameters$loc >= 0.2
}
if ("scale" %in% names(evd_parameters)) {
  valid_evd <- valid_evd & is.finite(evd_parameters$scale) & evd_parameters$scale > 0
}
evd_parameters[!valid_evd, ] <- NA_real_

return_specs <- data.frame(
  ARI_years = c(1, 40),
  AEP = 1 - exp(-1 / c(1, 40)),
  observed_name = c("Hs_ARI_1", "Hs_ARI_40"),
  stringsAsFactors = FALSE
)

return_levels <- lapply(seq_len(nrow(return_specs)), function(i) {
  loopevd::df_qevd(
    x = evd_parameters,
    evd_mod_str = "fgumbel",
    p = 1 - return_specs$AEP[i],
    interval = c(0, 20),
    cores = analysis_cores
  )
})

analysis_df <- data.frame(
  longitude = longitude,
  latitude = latitude,
  latitude_offset = latitude - params$reference_latitude,
  depth_m = average_depth_m,
  Hs_ARI_1 = return_levels[[1]],
  Hs_ARI_40 = return_levels[[2]]
)

# Fit an all-depth latitude benchmark plus shelf-depth models for both ARIs.
fit_results <- list()
fit_rows <- list()

calculate_model_metrics <- function(observed, predicted) {
  data.frame(
    R2 = 1 - sum((observed - predicted)^2, na.rm = TRUE) /
      sum((observed - mean(observed, na.rm = TRUE))^2, na.rm = TRUE),
    RMSE_m = sqrt(mean((observed - predicted)^2, na.rm = TRUE))
  )
}

for (i in seq_len(nrow(return_specs))) {
  observed_name <- return_specs$observed_name[i]
  latitude_only_name <- paste0(observed_name, "_latitude_only_model")
  depth_only_name <- paste0(observed_name, "_depth_only_model")
  predicted_name <- paste0(observed_name, "_depth_latitude_model")
  residual_name <- paste0(observed_name, "_depth_latitude_residual")

  latitude_domain <-
    is.finite(analysis_df$latitude_offset) &
    is.finite(analysis_df$depth_m) &
    analysis_df$depth_m > 0 &
    is.finite(analysis_df[[observed_name]]) &
    analysis_df[[observed_name]] > 0

  fit_domain <-
    is.finite(analysis_df$latitude_offset) &
    is.finite(analysis_df$depth_m) &
    analysis_df$depth_m >= params$minimum_depth_m &
    analysis_df$depth_m <= params$maximum_depth_m &
    is.finite(analysis_df[[observed_name]]) &
    analysis_df[[observed_name]] > 0

  latitude_data <- analysis_df[
    latitude_domain,
    c("latitude", "latitude_offset", observed_name)
  ]
  names(latitude_data)[3] <- "Hs_m"

  fit_data <- analysis_df[
    fit_domain,
    c("depth_m", "latitude", "latitude_offset", observed_name)
  ]
  names(fit_data)[4] <- "Hs_m"

  latitude_only_fit <- stats::lm(
    Hs_m ~ latitude_offset,
    data = latitude_data
  )
  depth_only_fit <- stats::lm(
    log10(Hs_m) ~ log10(depth_m),
    data = fit_data
  )

  depth_only_coefficients <- stats::coef(depth_only_fit)
  depth_scale_m <- 10^unname(depth_only_coefficients[1])
  depth_exponent <- unname(depth_only_coefficients[2])
  fit_data$depth_power <- fit_data$depth_m^depth_exponent

  depth_latitude_fit <- stats::lm(
    Hs_m ~ depth_power + latitude_offset,
    data = fit_data
  )

  latitude_only_prediction <- stats::predict(
    latitude_only_fit,
    newdata = latitude_data
  )
  depth_only_prediction <- 10^stats::predict(depth_only_fit, newdata = fit_data)
  combined_prediction <- stats::predict(depth_latitude_fit, newdata = fit_data)
  observed_latitude <- latitude_data$Hs_m
  observed_fit <- fit_data$Hs_m

  latitude_only_metrics <- calculate_model_metrics(
    observed_latitude,
    latitude_only_prediction
  )
  depth_only_metrics <- calculate_model_metrics(observed_fit, depth_only_prediction)
  combined_metrics <- calculate_model_metrics(observed_fit, combined_prediction)

  latitude_only_all <- rep(NA_real_, nrow(analysis_df))
  depth_only_all <- rep(NA_real_, nrow(analysis_df))
  predicted_all <- rep(NA_real_, nrow(analysis_df))
  latitude_only_all[latitude_domain] <- latitude_only_prediction
  depth_only_all[fit_domain] <- depth_only_prediction
  predicted_all[fit_domain] <- combined_prediction

  analysis_df[[latitude_only_name]] <- latitude_only_all
  analysis_df[[depth_only_name]] <- depth_only_all
  analysis_df[[predicted_name]] <- predicted_all
  analysis_df[[residual_name]] <- analysis_df[[observed_name]] - predicted_all

  latitude_only_coefficients <- stats::coef(latitude_only_fit)
  combined_coefficients <- stats::coef(depth_latitude_fit)

  fit_results[[paste0("ARI_", return_specs$ARI_years[i], "_latitude_only")]] <-
    latitude_only_fit
  fit_results[[paste0("ARI_", return_specs$ARI_years[i], "_depth_only")]] <-
    depth_only_fit
  fit_results[[paste0("ARI_", return_specs$ARI_years[i], "_depth_latitude")]] <-
    depth_latitude_fit

  fit_rows[[length(fit_rows) + 1]] <- data.frame(
    ARI_years = return_specs$ARI_years[i],
    AEP_percent = 100 * return_specs$AEP[i],
    Model = "Latitude only",
    intercept_m = unname(latitude_only_coefficients[1]),
    depth_scale_m = NA_real_,
    depth_exponent = NA_real_,
    latitude_coefficient = unname(latitude_only_coefficients[2]),
    R2 = latitude_only_metrics$R2,
    RMSE_m = latitude_only_metrics$RMSE_m,
    nodes = sum(latitude_domain),
    fit_depth_domain = "All valid positive depths"
  )

  fit_rows[[length(fit_rows) + 1]] <- data.frame(
    ARI_years = return_specs$ARI_years[i],
    AEP_percent = 100 * return_specs$AEP[i],
    Model = "Depth only",
    intercept_m = NA_real_,
    depth_scale_m = depth_scale_m,
    depth_exponent = depth_exponent,
    latitude_coefficient = NA_real_,
    R2 = depth_only_metrics$R2,
    RMSE_m = depth_only_metrics$RMSE_m,
    nodes = sum(fit_domain),
    fit_depth_domain = paste0(
      params$minimum_depth_m, "–", params$maximum_depth_m, " m"
    )
  )

  fit_rows[[length(fit_rows) + 1]] <- data.frame(
    ARI_years = return_specs$ARI_years[i],
    AEP_percent = 100 * return_specs$AEP[i],
    Model = "Depth + latitude",
    intercept_m = unname(combined_coefficients[1]),
    depth_scale_m = unname(combined_coefficients[2]),
    depth_exponent = depth_exponent,
    latitude_coefficient = unname(combined_coefficients[3]),
    R2 = combined_metrics$R2,
    RMSE_m = combined_metrics$RMSE_m,
    nodes = sum(fit_domain),
    fit_depth_domain = paste0(
      params$minimum_depth_m, "–", params$maximum_depth_m, " m"
    )
  )
}

fit_summary <- do.call(rbind, fit_rows)
latitude_only_summary <- fit_summary[fit_summary$Model == "Latitude only", ]
latitude_only_summary <- latitude_only_summary[
  match(return_specs$ARI_years, latitude_only_summary$ARI_years),
]
primary_fit_summary <- fit_summary[fit_summary$Model == "Depth + latitude", ]
primary_fit_summary <- primary_fit_summary[
  match(return_specs$ARI_years, primary_fit_summary$ARI_years),
]
```

# Results: how much do depth and latitude explain?

```{r parameter-table, echo=FALSE, results='asis'}
depth_only_summary <- fit_summary[fit_summary$Model == "Depth only", ]
depth_only_summary <- depth_only_summary[
  match(return_specs$ARI_years, depth_only_summary$ARI_years),
]

coefficient_table <- data.frame(
  `ARI (years)` = fit_summary$ARI_years,
  `AEP (%)` = sprintf("%.2f", fit_summary$AEP_percent),
  `Model` = fit_summary$Model,
  `Additive intercept (m)` = ifelse(
    is.na(fit_summary$intercept_m),
    "—",
    sprintf("%.3f", fit_summary$intercept_m)
  ),
  `Depth scale C<br>at h = 1 m (m)` = ifelse(
    is.na(fit_summary$depth_scale_m),
    "—",
    sprintf("%.3f", fit_summary$depth_scale_m)
  ),
  `Depth exponent<br>(power on h)` = ifelse(
    is.na(fit_summary$depth_exponent),
    "—",
    sprintf("%.3f", fit_summary$depth_exponent)
  ),
  `Latitude coefficient<br>(m per degree)` = ifelse(
    is.na(fit_summary$latitude_coefficient),
    "—",
    sprintf("%.4f", fit_summary$latitude_coefficient)
  ),
  check.names = FALSE
)

performance_table <- data.frame(
  `ARI (years)` = fit_summary$ARI_years,
  `Model` = fit_summary$Model,
  `R-squared` = sprintf("%.3f", fit_summary$R2),
  `RMSE (m)` = sprintf("%.3f", fit_summary$RMSE_m),
  `Fit-depth domain` = fit_summary$fit_depth_domain,
  `Model nodes` = format(fit_summary$nodes, big.mark = ",", scientific = FALSE),
  check.names = FALSE
)

knitr::kable(
  coefficient_table,
  align = c("r", "r", "l", "r", "r", "r", "r"),
  escape = FALSE,
  caption = paste0(
    "Power-law depth and original-scale latitude coefficients. The additive ",
    "intercept and latitude effect are evaluated relative to ",
    abs(params$reference_latitude), "°S."
  )
)

cat("\n\n")

knitr::kable(
  performance_table,
  align = c("r", "l", "r", "r", "l", "r"),
  caption = paste0(
    "Model performance. The latitude-only benchmark uses all valid positive ",
    "depths; the other models use the stated shelf-depth range."
  )
)
```

The latitude-only relationship and the latitude term in the combined model remain linear in $H_s$ metres. Only the depth exponent is estimated from $\log_{10}(H_s)$ versus $\log_{10}(h)$. To keep the performance statistics comparable, every reported $R^2$ and RMSE is calculated after prediction on the original $H_s$ scale in metres. The all-depth latitude-only model uses the exact 1-year and 40-year ARI $H_s$ at every node with finite latitude, positive finite depth and positive finite return-level wave height. Its $R^2$ values are `r sprintf("%.3f", latitude_only_summary$R2[1])` for the 1-year extreme and `r sprintf("%.3f", latitude_only_summary$R2[2])` for the 40-year extreme. Latitude alone can still have a low national $R^2$ because it does not represent depth, exposure, sheltering, fetch or storm type. These values also describe a broader depth domain and are therefore a benchmark rather than a like-for-like test against the two shelf-depth models.

For the 1-year extreme, adding latitude changes $R^2$ from `r sprintf("%.3f", depth_only_summary$R2[1])` to `r sprintf("%.3f", primary_fit_summary$R2[1])`; for the 40-year extreme it changes from `r sprintf("%.3f", depth_only_summary$R2[2])` to `r sprintf("%.3f", primary_fit_summary$R2[2])`. This comparison uses identical shelf-depth nodes. Any improvement shows that the broad north–south storm climate contains information not represented by depth alone, but it does not make latitude a complete description of storm exposure.

The fitted depth exponent is not a universal breaker coefficient. It is an empirical summary of the national CCHaPS relationship, while the combined model adds a linear latitude correction in metres; shelf environments at the same depth and latitude can still differ greatly in fetch, coastline geometry, sheltering and incoming swell.

<div class="plain-language">
<strong>In plain language</strong>
The depth-only $R^2$ increases from `r sprintf("%.3f", depth_only_summary$R2[1])` for the 1-year ARI to `r sprintf("%.3f", depth_only_summary$R2[2])` for the 40-year ARI, while the depth-and-latitude $R^2$ increases from `r sprintf("%.3f", primary_fit_summary$R2[1])` to `r sprintf("%.3f", primary_fit_summary$R2[2])`. This stronger relationship for the 40-year event is consistent with the larger waves approaching the depth limit across a wider range of shelf depths, so water depth becomes a stronger control. It is supporting evidence rather than proof that every mapped 40-year wave is actively breaking.

The latitude-only $R^2$ is much smaller in this Australian node-by-node analysis than in the global gridded analysis. One likely contributor is the CCHaPS mesh: it contains many closely spaced nodes in shallow coastal waters and fewer, more widely spaced nodes in deep water. The regression gives every node equal weight, so spatially variable shallow-water environments receive more weight than their geographic area would imply. Depth, exposure, sheltering, fetch and storm type also contribute to the remaining variation.
</div>

```{r depth-scatter, echo=FALSE, fig.width=11, fig.height=6.4, fig.cap="Figure 1. CCHaPS extreme significant wave height against average water depth for the 1-year ARI (left) and 40-year ARI (right). Sampled CCHaPS nodes are coloured by latitude. Both axes are logarithmic, matching the original power-law depth diagnostic; the wave-height range is restricted to 0.2–12 m."}
set.seed(20260818)
old_par <- par(no.readonly = TRUE)
layout(
  matrix(c(1, 2, 3, 3), nrow = 2, ncol = 2, byrow = TRUE),
  heights = c(1, 0.16)
)
par(
  mar = c(4.8, 5.1, 3.6, 1.2),
  oma = c(0, 0, 0, 0),
  cex.axis = 1.02,
  cex.lab = 1.18,
  cex.main = 1.22
)

latitude_limits <- range(analysis_df$latitude, na.rm = TRUE)
latitude_palette <- grDevices::colorRampPalette(c(
  "#313695", "#74add1", "#ffffbf", "#f46d43", "#a50026"
))(256)
latitude_to_colour <- function(latitude_value, alpha = 1) {
  scaled_latitude <-
    (latitude_value - latitude_limits[1]) / diff(latitude_limits)
  colour_index <- pmax(
    1,
    pmin(256, 1 + floor(255 * scaled_latitude))
  )
  grDevices::adjustcolor(latitude_palette[colour_index], alpha.f = alpha)
}

for (i in seq_len(nrow(return_specs))) {
  observed_name <- return_specs$observed_name[i]
  valid <-
    is.finite(analysis_df$latitude) &
    is.finite(analysis_df$depth_m) &
    analysis_df$depth_m >= params$minimum_depth_m &
    analysis_df$depth_m <= params$maximum_depth_m &
    is.finite(analysis_df[[observed_name]]) &
    analysis_df[[observed_name]] > 0

  valid_indices <- which(valid)
  sample_indices <- if (length(valid_indices) > 120000) {
    sample(valid_indices, 120000)
  } else {
    valid_indices
  }

  plot(
    analysis_df$depth_m[sample_indices],
    analysis_df[[observed_name]][sample_indices],
    log = "xy",
    pch = 16,
    cex = 0.23,
    col = latitude_to_colour(
      analysis_df$latitude[sample_indices],
      alpha = 0.16
    ),
    xlim = c(params$minimum_depth_m, params$maximum_depth_m),
    ylim = c(0.2, 12),
    xlab = "Average water depth, h (m)",
    ylab = "Extreme significant wave height, Hs (m)",
    main = paste0("ARI = ", return_specs$ARI_years[i], " year",
                  if (return_specs$ARI_years[i] > 1) "s" else "")
  )
  grid(col = grDevices::adjustcolor("grey70", 0.5))
}

par(mar = c(0.4, 5.1, 0.4, 1.2))
plot.new()
plot.window(
  xlim = latitude_limits,
  ylim = c(0, 1),
  xaxs = "i",
  yaxs = "i"
)
latitude_key_edges <- seq(
  latitude_limits[1],
  latitude_limits[2],
  length.out = length(latitude_palette) + 1
)
rect(
  xleft = latitude_key_edges[-length(latitude_key_edges)],
  ybottom = 0.40,
  xright = latitude_key_edges[-1],
  ytop = 0.68,
  col = latitude_palette,
  border = NA
)
rect(
  xleft = latitude_limits[1],
  ybottom = 0.40,
  xright = latitude_limits[2],
  ytop = 0.68,
  border = "grey40"
)
latitude_ticks <- pretty(latitude_limits, n = 5)
latitude_ticks <- latitude_ticks[
  latitude_ticks >= latitude_limits[1] &
    latitude_ticks <= latitude_limits[2]
]
latitude_tick_labels <- ifelse(
  latitude_ticks < 0,
  paste0(abs(latitude_ticks), "°S"),
  ifelse(latitude_ticks > 0, paste0(latitude_ticks, "°N"), "0°")
)
axis(
  1,
  at = latitude_ticks,
  labels = latitude_tick_labels,
  pos = 0.40,
  cex.axis = 1.05,
  tcl = -0.24,
  lwd = 0,
  lwd.ticks = 0.8,
  padj = 0.2
)
text(
  x = mean(latitude_limits),
  y = 0.98,
  labels = "Latitude of CCHaPS node",
  adj = c(0.5, 1),
  cex = 1.16,
  font = 2
)

layout(1)
par(old_par)
```

<div class="plain-language">
<strong>In plain language</strong>
The upper edge of the point cloud rises with depth because deeper water can support larger waves before depth-induced breaking becomes important. Colouring the nodes by latitude shows the broad north–south climate pattern without hiding the substantial overlap: a shallow but exposed coast can still be rougher than a deeper but sheltered gulf.
</div>

## How large are the waves compared with the available depth?

The ratio $H_s/h$ is useful for comparing the event height with the water beneath it. It is **not** a universal breaking threshold, but larger ratios indicate that the modelled sea state occupies a larger share of the available depth and is more likely to be affected by shallow-water processes.

```{r depth-ratio, echo=FALSE, fig.width=9.5, fig.height=5.4, fig.cap="Figure 2. Median significant-wave-height-to-depth ratio within logarithmically spaced depth bands. Both axes are logarithmic. The ratio generally falls as water becomes deeper. The 40-year event occupies more of the available depth than the 1-year event, so depth limitation is expected to matter more strongly for the rarer, larger waves."}
depth_edges <- 10^seq(
  log10(params$minimum_depth_m),
  log10(params$maximum_depth_m),
  length.out = 30
)
depth_centres <- sqrt(depth_edges[-1] * depth_edges[-length(depth_edges)])

ratio_medians <- matrix(
  NA_real_,
  nrow = length(depth_centres),
  ncol = nrow(return_specs)
)

for (i in seq_len(nrow(return_specs))) {
  observed <- analysis_df[[return_specs$observed_name[i]]]
  valid <-
    is.finite(analysis_df$depth_m) &
    analysis_df$depth_m >= params$minimum_depth_m &
    analysis_df$depth_m <= params$maximum_depth_m &
    is.finite(observed) & observed > 0

  depth_group <- cut(
    analysis_df$depth_m[valid],
    breaks = depth_edges,
    include.lowest = TRUE,
    labels = FALSE
  )
  ratio <- observed[valid] / analysis_df$depth_m[valid]

  ratio_medians[, i] <- vapply(seq_along(depth_centres), function(bin_number) {
    stats::median(ratio[depth_group == bin_number], na.rm = TRUE)
  }, numeric(1))
}

matplot(
  depth_centres,
  ratio_medians,
  type = "l",
  log = "xy",
  lty = 1,
  lwd = 3,
  col = c("#238b45", "#cb181d"),
  xlab = "Average water depth, h (m)",
  ylab = "Median Hs / h",
  main = "How much of the available depth does the extreme wave occupy?"
)
grid(col = grDevices::adjustcolor("grey70", 0.5))
legend(
  "topright",
  legend = paste0("ARI = ", return_specs$ARI_years, c(" year", " years")),
  col = c("#238b45", "#cb181d"),
  lwd = 3,
  bty = "n"
)
```

# Map comparison

The CCHaPS calculations are supplied on an unstructured mesh rather than a regular raster. For plotting only, each model node is assigned to a regular `r params$map_resolution_degrees`° longitude–latitude cell; where several nodes fall inside one cell, their values are averaged. Empty display cells are filled from their immediate neighbours for `r params$map_gap_fill_passes` short passes, then the rasters are clipped back to the fitted shelf-depth range and masked over land. This rasterisation closes small visual gaps and provides a consistent map grid, but it does not change the regressions, which are fitted to the original unstructured nodes.

```{r prepare-map-rasters, echo=TRUE, results='hide'}
fit_map_domain <-
  is.finite(analysis_df$depth_m) &
  analysis_df$depth_m >= params$minimum_depth_m &
  analysis_df$depth_m <= params$maximum_depth_m

map_fields <- c(
  "depth_m",
  "Hs_ARI_1", "Hs_ARI_40",
  "Hs_ARI_1_depth_latitude_model", "Hs_ARI_40_depth_latitude_model",
  "Hs_ARI_1_depth_latitude_residual", "Hs_ARI_40_depth_latitude_residual"
)

map_data <- analysis_df[
  fit_map_domain,
  c("longitude", "latitude", map_fields)
]

map_points <- terra::vect(
  map_data,
  geom = c("longitude", "latitude"),
  crs = "EPSG:4326"
)

australia_template <- terra::rast(
  xmin = 110,
  xmax = 155,
  ymin = -45,
  ymax = -5,
  resolution = params$map_resolution_degrees,
  crs = "EPSG:4326"
)

raw_map_raster_list <- lapply(map_fields, function(field_name) {
  terra::rasterize(
    map_points,
    australia_template,
    field = field_name,
    fun = "mean",
    na.rm = TRUE
  )
})

raw_map_rasters <- terra::rast(raw_map_raster_list)
names(raw_map_rasters) <- map_fields

coastline <- terra::vect(input_files$coastline)
coastline <- terra::as.polygons(coastline)
coastline <- terra::project(coastline, "EPSG:4326")

fill_small_display_gaps <- function(raster_layer, passes = 2) {
  output <- raster_layer
  if (passes <= 0) return(output)

  for (pass_number in seq_len(passes)) {
    neighbour_mean <- terra::focal(
      output,
      w = 3,
      fun = "mean",
      na.rm = TRUE,
      na.policy = "only",
      fillvalue = NA
    )
    output <- terra::cover(output, neighbour_mean)
  }
  output
}

# Fill only the display rasters; the statistical models continue to use nodes.
filled_depth <- fill_small_display_gaps(
  raw_map_rasters[["depth_m"]],
  params$map_gap_fill_passes
)

fitted_shelf_mask <- terra::ifel(
  filled_depth >= params$minimum_depth_m &
    filled_depth <= params$maximum_depth_m,
  1,
  NA
)

coastline$land_mask_value <- 1
land_mask <- terra::rasterize(
  coastline,
  australia_template,
  field = "land_mask_value",
  background = NA
)

map_raster_list <- lapply(seq_len(terra::nlyr(raw_map_rasters)), function(i) {
  filled_layer <- fill_small_display_gaps(
    raw_map_rasters[[i]],
    params$map_gap_fill_passes
  )
  filled_layer <- terra::mask(filled_layer, fitted_shelf_mask)
  terra::mask(filled_layer, land_mask, inverse = TRUE)
})

map_rasters <- terra::rast(map_raster_list)
names(map_rasters) <- map_fields

plot_map_panel <- function(
  raster_layer,
  title,
  colours,
  breaks,
  show_legend = TRUE,
  title_cex = 1.15
) {
  if (show_legend) {
    terra::plot(
      raster_layer,
      col = colours,
      breaks = breaks,
      axes = FALSE,
      main = title,
      cex.main = title_cex,
      plg = list(cex = 0.8, title = "m")
    )
  } else {
    terra::plot(
      raster_layer,
      col = colours,
      breaks = breaks,
      axes = FALSE,
      main = title,
      cex.main = title_cex,
      legend = FALSE
    )
  }
  terra::plot(
    terra::crop(coastline, terra::ext(raster_layer)),
    add = TRUE,
    col = "grey88",
    border = "grey55",
    lwd = 0.25
  )
  box(col = "grey45")
}

draw_shared_colour_bar <- function(
  breaks,
  colours,
  label,
  tick_positions,
  tick_labels = tick_positions,
  axis_cex = 0.90,
  label_cex = 1.06
) {
  plot.new()
  plot.window(
    xlim = range(breaks) + c(-1, 1) * diff(range(breaks)) * 0.025,
    ylim = c(0, 1),
    xaxs = "i",
    yaxs = "i"
  )
  rect(
    xleft = breaks[-length(breaks)],
    ybottom = 0.46,
    xright = breaks[-1],
    ytop = 0.72,
    col = colours,
    border = NA
  )
  rect(
    xleft = min(breaks),
    ybottom = 0.46,
    xright = max(breaks),
    ytop = 0.72,
    border = "grey45"
  )
  axis(
    1,
    at = tick_positions,
    labels = tick_labels,
    pos = 0.46,
    cex.axis = axis_cex,
    tcl = -0.24,
    lwd = 0,
    lwd.ticks = 0.8,
    padj = 0.2
  )
  text(
    x = mean(range(breaks)),
    y = 0.98,
    labels = label,
    adj = c(0.5, 1),
    cex = label_cex,
    font = 2
  )
}
```

```{r fitted-depth-map, echo=FALSE, fig.width=10.5, fig.height=7.8, fig.cap="Figure 3. Average CCHaPS water depth across the fitted domain. Pale colours show shallow banks and inner shelves; darker blues show the deeper part of the fitted continental shelf. Areas outside the selected depth range are excluded from the analysis. This map has no logarithmic axis or colour transformation: depth is shown in linear 5 m intervals."}
depth_breaks <- seq(0, params$maximum_depth_m, by = 5)
depth_colours <- rev(
  grDevices::hcl.colors(length(depth_breaks) - 1, "YlGnBu")
)

old_par <- par(no.readonly = TRUE)
layout(matrix(c(1, 2), nrow = 2), heights = c(1, 0.16))
par(mar = c(1.0, 1.0, 3.0, 1.0), oma = c(0, 0, 0, 0))
plot_map_panel(
  map_rasters[["depth_m"]],
  "Average water depth used in the fit",
  depth_colours,
  depth_breaks,
  show_legend = FALSE,
  title_cex = 1.45
)

par(mar = c(0.2, 4.2, 0.2, 4.2))
draw_shared_colour_bar(
  depth_breaks,
  depth_colours,
  "Average water depth (m)",
  tick_positions = depth_breaks,
  axis_cex = 1.08,
  label_cex = 1.25
)

layout(1)
par(old_par)
```

```{r six-panel-map, echo=FALSE, fig.width=12, fig.height=15.5, fig.cap="Figure 4. Extreme significant wave height across the fitted shelf-depth range. Columns compare the 1-year ARI (left) with the 40-year ARI (right). The top row is the full CCHaPS estimate, the middle row is the depth-and-latitude model, and the bottom row is the residual (CCHaPS minus the simple model). The two shared colour bars apply to the four wave-height panels and two residual panels, respectively. The mapped wave heights and residuals are untransformed values in metres on linear colour scales; no logarithmic axis or colour transformation is used."}
old_par <- par(no.readonly = TRUE)
layout(
  matrix(
    c(1, 2, 3, 4, 5, 6, 7, 7, 8, 8),
    nrow = 5,
    ncol = 2,
    byrow = TRUE
  ),
  heights = c(1, 1, 1, 0.24, 0.24)
)
par(mar = c(1.0, 1.0, 3.4, 1.0), oma = c(0, 0, 0, 0))

plot_map_panel(
  map_rasters[["Hs_ARI_1"]],
  "CCHaPS: ARI = 1 year",
  colzRamp(length(wave_breaks) - 1),
  wave_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)
plot_map_panel(
  map_rasters[["Hs_ARI_40"]],
  "CCHaPS: ARI = 40 years",
  colzRamp(length(wave_breaks) - 1),
  wave_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)
plot_map_panel(
  map_rasters[["Hs_ARI_1_depth_latitude_model"]],
  "Depth + latitude: ARI = 1 year",
  colzRamp(length(wave_breaks) - 1),
  wave_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)
plot_map_panel(
  map_rasters[["Hs_ARI_40_depth_latitude_model"]],
  "Depth + latitude: ARI = 40 years",
  colzRamp(length(wave_breaks) - 1),
  wave_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)
plot_map_panel(
  terra::clamp(
    map_rasters[["Hs_ARI_1_depth_latitude_residual"]],
    lower = -params$residual_display_max_m,
    upper = params$residual_display_max_m,
    values = TRUE
  ),
  "Residual: ARI = 1 year",
  bwr(length(residual_breaks) - 1),
  residual_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)
plot_map_panel(
  terra::clamp(
    map_rasters[["Hs_ARI_40_depth_latitude_residual"]],
    lower = -params$residual_display_max_m,
    upper = params$residual_display_max_m,
    values = TRUE
  ),
  "Residual: ARI = 40 years",
  bwr(length(residual_breaks) - 1),
  residual_breaks,
  show_legend = FALSE,
  title_cex = 1.65
)

par(mar = c(0.2, 4.2, 0.2, 4.2))
draw_shared_colour_bar(
  wave_breaks,
  colzRamp(length(wave_breaks) - 1),
  "Extreme significant wave height, Hs (m)",
  tick_positions = wave_breaks,
  axis_cex = 1.26,
  label_cex = 1.42
)

par(mar = c(0.2, 4.2, 0.2, 4.2))
draw_shared_colour_bar(
  residual_breaks,
  bwr(length(residual_breaks) - 1),
  "Residual: CCHaPS minus depth-and-latitude model (m)",
  tick_positions = residual_breaks,
  tick_labels = ifelse(
    abs(residual_breaks - round(residual_breaks)) < 1e-8,
    sprintf("%.0f", residual_breaks),
    ""
  ),
  axis_cex = 1.26,
  label_cex = 1.42
)

layout(1)
par(old_par)
```

The wave-height colour scale runs from 0 to `r params$wave_display_max_m` m in 1 m increments: blues show values below 5 m, yellows and reds show values from 5 to 11 m, and browns, purples and pinks show values above 11 m. The residual display runs from -`r params$residual_display_max_m` to `r params$residual_display_max_m` m in discrete `r params$residual_display_increment_m` m increments. Blue means CCHaPS is lower than the depth-and-latitude estimate, white marks the transition around zero, and red or purple means CCHaPS is higher. Values outside the displayed residual range are clipped to its end colours.

<div class="plain-language">
<strong>In plain language</strong>
Depth limitation is more visible for the rarer 40-year waves because a larger incoming event comes closer to the maximum size that shallow water can support. The remaining regional patterns are a reminder that the incoming storm climate, tropical cyclones, swell pathways, fetch, sheltering and shelf shape determine how much wave energy reaches each depth.
</div>

# Explore the layers interactively

Use the radio buttons at bottom left to switch between the CCHaPS estimate, depth-and-latitude model and residual for each ARI. The grey background is the Esri World Light Gray Canvas basemap. Small gaps in the regular display grid have been filled from neighbouring cells, while the fitted shelf-depth and land masks are retained.

```{r interactive-map, echo=FALSE}
wave_palette <- leaflet::colorBin(
  palette = colzRamp(length(wave_breaks) - 1),
  bins = wave_breaks,
  domain = c(0, params$wave_display_max_m),
  na.color = "transparent",
  pretty = FALSE,
  right = FALSE
)

residual_palette <- leaflet::colorBin(
  palette = bwr(length(residual_breaks) - 1),
  bins = residual_breaks,
  domain = c(-params$residual_display_max_m, params$residual_display_max_m),
  na.color = "transparent",
  pretty = FALSE,
  right = FALSE
)

interactive_groups <- c(
  "CCHaPS — ARI 1 year",
  "CCHaPS — ARI 40 years",
  "Depth + latitude — ARI 1 year",
  "Depth + latitude — ARI 40 years",
  "Residual — ARI 1 year",
  "Residual — ARI 40 years"
)

wave_display <- lapply(
  c(
    "Hs_ARI_1", "Hs_ARI_40",
    "Hs_ARI_1_depth_latitude_model", "Hs_ARI_40_depth_latitude_model"
  ),
  function(layer_name) {
    terra::clamp(
      map_rasters[[layer_name]],
      lower = 0,
      upper = params$wave_display_max_m,
      values = TRUE
    )
  }
)

residual_display <- lapply(
  c(
    "Hs_ARI_1_depth_latitude_residual",
    "Hs_ARI_40_depth_latitude_residual"
  ),
  function(layer_name) {
    terra::clamp(
      map_rasters[[layer_name]],
      lower = -params$residual_display_max_m,
      upper = params$residual_display_max_m,
      values = TRUE
    )
  }
)

depth_map <- leaflet::leaflet(
  options = leaflet::leafletOptions(minZoom = 3, preferCanvas = TRUE)
) |>
  leaflet::addProviderTiles(
    leaflet::providers$Esri.WorldGrayCanvas,
    group = "Light Gray Canvas"
  )

for (i in seq_along(wave_display)) {
  depth_map <- leaflet::addRasterImage(
    depth_map,
    wave_display[[i]],
    colors = wave_palette,
    opacity = 0.83,
    group = interactive_groups[i],
    project = TRUE,
    maxBytes = 64 * 1024 * 1024
  )
}

for (i in seq_along(residual_display)) {
  depth_map <- leaflet::addRasterImage(
    depth_map,
    residual_display[[i]],
    colors = residual_palette,
    opacity = 0.83,
    group = interactive_groups[i + 4],
    project = TRUE,
    maxBytes = 64 * 1024 * 1024
  )
}

depth_map |>
  leaflet::hideGroup(interactive_groups[-1]) |>
  leaflet::addLayersControl(
    baseGroups = interactive_groups,
    position = "bottomright",
    options = leaflet::layersControlOptions(collapsed = FALSE)
  ) |>
  leaflet::addLegend(
    position = "topright",
    pal = wave_palette,
    values = wave_breaks,
    title = htmltools::HTML("Extreme H<sub>s</sub> (m)"),
    opacity = 0.9
  ) |>
  leaflet::addLegend(
    position = "bottomleft",
    pal = residual_palette,
    values = residual_breaks,
    title = htmltools::HTML("CCHaPS − depth +<br>latitude (m)"),
    opacity = 0.9
  ) |>
  leaflet::fitBounds(110, -45, 155, -5)
```


# What does the depth-and-latitude model miss?

A red area of positive residual does not mean the simple calculation is wrong. It means the full CCHaPS wave climate is more energetic than expected from depth and the broad latitude effect. Likely contributors include exposure to Southern Ocean swell, tropical-cyclone or east-coast-low waves, long fetch, coastline orientation, wave focusing and connections to deeper water.

A blue area of negative residuals means the full model is lower than the depth-and-latitude estimate. This may occur inside sheltered gulfs, behind islands or reefs, across short fetches, or where bottom friction and repeated breaking remove energy before waves reach the model node.

<div class="plain-language">
<strong>In plain language</strong>
Blue areas are clearly sheltered or fetch-limited environments where CCHaPS estimates smaller extreme waves than the depth-and-latitude model. Red areas are places where CCHaPS estimates waves that are much higher than water depth and latitude alone would predict. The red pattern is more evident for the 40-year ARI and typically occurs in deeper or more exposed water, including coasts exposed to Southern Ocean swell, parts of the North West Shelf and waters seaward of the Great Barrier Reef. White areas are close to zero residual, where the simple model and CCHaPS give similar wave heights. These colours describe the direction and size of the difference; they are not a measure of confidence or proof that either estimate is wrong.
</div>

Future analysis could separate the depth effect more cleanly by including offshore wave exposure, tropical-cyclone tracks, distance from the coast, shelf width, bathymetric slope, coastline orientation and fetch. A particularly useful next step would compare nodes with similar offshore wave forcing and latitude but different local depth.

# Where does the depth-limited wave energy go?

Depth-induced breaking removes energy from the short-period wind-wave spectrum, but “dissipated” does not mean that all of the lost energy reappears in one other wave component. Much of it becomes breaking-generated turbulence and is ultimately converted to heat. At the same time, the shoreward reduction in wave momentum flux—usually described through radiation-stress gradients—changes mean water level through wave setup and forces cross-shore, rip and longshore currents. The foundations of these momentum transfers are described by [Longuet-Higgins and Stewart (1962)](https://doi.org/10.1017/S0022112062000877), with a classic longshore-current application in [Longuet-Higgins (1970)](https://doi.org/10.1029/JC075i033p06778).

Wave groups also create slow variations in wave energy and radiation stress. Bound long waves accompany the groups outside the surf zone. As the groups shoal and break, the moving breakpoint and group-varying breaking can generate or release free infragravity waves, observed as **surf beat**. [Symonds, Huntley and Bowen (1982)](https://doi.org/10.1029/JC087iC01p00492) describe this moving-breakpoint mechanism. It is an important pathway for transferring and generating low-frequency motion, but it does not imply that all short-wave energy lost during breaking becomes infragravity-wave energy.

Breaking turbulence, wave orbital motion and wave-driven currents can also mobilise the seabed. Sediment may move as bedload close to the bed or as suspended load in the water column. Spatial differences in that transport can build or erode bars, rip channels, ripples and other bedforms, altering the bathymetry that controls later waves. Models such as [XBeach](https://doi.org/10.1016/j.coastaleng.2009.08.006) couple wave groups and infragravity motion with currents, sediment transport and bed evolution for storm-impact studies. Currents and sediment transport should not, however, be treated as simple energy “bins”: the partition depends on wave period and angle, beach slope, water depth, sediment properties and the evolving morphology, while most broken-wave energy still cascades through turbulence to heat.

<div class="plain-language">
<strong>In plain language</strong>
A breaking storm wave does not simply turn into a long wave, a current or moving sand. Breaking removes energy from the ordinary wind waves, creates turbulence and changes the force that the waves exert on the water. Those changes can raise the mean water level, drive longshore and rip currents, generate slow surf-beat or infragravity waves, and move large quantities of bed and suspended sediment. These linked responses can reshape bars and channels, while much of the original wave energy is eventually lost as heat.
</div>

CCHaPS is a national, phase-averaged circulation-and-wave model. It represents bulk wave transformation and depth-induced breaking, but it is not intended to resolve each individual breaker, beach-scale infragravity motion or storm-driven bedform evolution. A future local analysis could combine CCHaPS offshore boundary conditions with observations and a finer surf-zone or morphodynamic model to track these energy and momentum pathways explicitly.

# Looking ahead: sea-level rise and changing storms

Sea-level rise increases the water depth over reefs, banks and shallow shelves. Where waves are currently depth limited, additional depth may allow larger waves to travel farther shoreward before breaking. The response will not be a simple one-for-one increase because wave period, bottom friction, currents, changing storm tracks and changing offshore wave climate also matter.

This historical relationship should therefore not be assumed to remain fixed. A future version could repeat the analysis with the CCHaPS climate projections and separate the effects of changing water depth from changes in wind, storms and incoming wave spectra.

A further analysis could test the conceptual connection with the [Dean equilibrium beach profile](https://journals.flvc.org/jcr/article/view/78405), $h(x)=Ax^{2/3}$. If a local limiting wave height is approximated as $H_{\mathrm{lim}}\approx\gamma h$, then a Dean-like beach profile implies $H_{\mathrm{lim}}(x)\approx\gamma A x^{2/3}$. This would describe how a depth-limited wave ceiling changes with cross-shore distance; it would not imply that the national fitted depth exponent must equal $2/3$. A direct test would require shore-normal beach transects, distance from the coast, local sediment information or a fitted $A$, and comparable offshore wave forcing. The present nodes span shelves, gulfs, banks, reefs and non-equilibrium settings, so this report does not estimate $A$ or test the Dean exponent. The formulation and its applications are described further by [Dean (1991)](https://journals.flvc.org/jcr/article/download/78405/75816/).

# Limits and reproducibility notes

- The depth-and-latitude model describes a national association; it does not isolate either variable as a causal effect.
- The latitude-only benchmark uses all valid positive water depths. The depth-only and depth-and-latitude models are restricted to `r params$minimum_depth_m`–`r params$maximum_depth_m` m and should not be extrapolated beyond those limits.
- The latitude-only model and the latitude term in the combined model are linear in $H_s$ metres. The depth exponent is estimated from the original $\log_{10}(H_s)$–$\log_{10}(h)$ relationship, and all performance metrics are evaluated in metres.
- The Dean profile connection is conceptual at this scale. This analysis does not estimate a beach-profile parameter $A$, use shore-normal distance, or test the $2/3$ profile exponent.
- The analysis uses Gumbel fits to annual maxima for consistency with the attached depth workflow. It does not compare alternative extreme-value distributions.
- A 40-year hindcast contains limited information about a 40-year ARI, particularly in tropical-cyclone regions. The uncertainty of the return level should be considered in any applied use.
- CCHaPS is a national model. Individual beaches, reefs, bars and harbour entrances may require substantially finer bathymetry and local modelling.
- Static and interactive maps rasterise the unstructured model nodes onto a regular `r params$map_resolution_degrees`° display grid and fill only small neighbouring-cell gaps; model fitting uses the original nodes.
- The 0.0125° display grid contains sixteen times as many cells as the original 0.05° grid, increasing memory use, knitting time and HTML size without adding new model nodes.
- Display gap filling can smooth narrow channels or shelf-edge details and should not be used as a substitute for the original node values.
- The report describes offshore or shelf $H_s$, not the height of a breaking surf wave at a beach and not the largest individual wave.

## Input files

| Purpose | Expected file beneath `params$datadir` |
|:--|:--|
| Bathymetry and mesh coordinates | `depth_cchaps_hindcast_BARRA-R2_WHACS_ERA5_fx.nc` |
| Annual mean sea-surface height | `timeseries/zos_cchaps_hindcast_BARRA-R2_WHACS_ERA5_yr_mean_198101010000-202012310000.nc` |
| Gumbel fits to annual-maximum $H_s$ | `EVA/hs_cchaps_hindcast_BARRA-R2_WHACS_ERA5_yr_max_198101010000-202012310000_fgumbel.nc` |
| Australian coastline polygons | `LGA_2023_AUST_GDA2020_coastline` |

## Changes made to the exploratory script

The report keeps the original CCHaPS inputs, mean-water-level depth correction, Gumbel return-level calculation and $H_s/h$ diagnostic. It also:

- calculates exact 1-year and 40-year ARIs rather than using the nearest pre-computed AEP columns;
- checks that bathymetry and extreme-value files use the same node order;
- fits and reports both return intervals consistently;
- adds an all-depth, original-scale latitude-only benchmark and an additive linear latitude term to the shelf-depth model;
- restores the original log–log depth fit, keeps the latitude effect additive in $H_s$ metres and reports original-scale $R^2$ and RMSE;
- colours the depth-versus-wave-height point cloud by latitude;
- masks maps to the fitted depth range;
- replaces a national Voronoi operation with regular-grid averaging and limited gap filling for more practical HTML maps;
- uses two shared colour bars instead of six compressed legends in the tiled map;
- displays residuals from -5 to 5 m in discrete 0.5 m colours and places colour-key titles above their bars;
- fixes the inconsistent `coastline`/`coastline_oz` object name; and
- removes exploratory AIC, alternative-distribution and confidence-interval calculations that were not used by the final model.

## Data sources

The fitted Gumbel parameters and return-level information used here come from the CSIRO Data Access Portal collection [*Extreme Value Analysis of CCHaPS*](https://data.csiro.au/collection/csiro%3A65867). The bathymetry, mesh coordinates and mean water levels are from the associated CCHaPS hindcast files listed above.

For further reading about the CCHaPS model design, implementation and validation, see:

Hernaman, V., et al. (2026). *The Australian Climate Service Coupled circulation-wave Coastal Hazards Prediction System (CCHaPS): Development, implementation, and validation*. **Ocean Modelling, 202**, 102731. [https://doi.org/10.1016/j.ocemod.2026.102731](https://doi.org/10.1016/j.ocemod.2026.102731)

<script>
(function () {
  function initialiseLightbox() {
    var overlay = document.getElementById('figure-lightbox');
    if (!overlay) return;
    var enlarged = overlay.querySelector('img');
    var closeButton = overlay.querySelector('.lightbox-close');

    function closeLightbox() {
      overlay.classList.remove('open');
      enlarged.src = '';
    }

    document.querySelectorAll('.main-container img').forEach(function (img) {
      if (img.closest('.leaflet-container') || img.closest('#figure-lightbox')) return;
      img.setAttribute('tabindex', '0');
      img.setAttribute('role', 'button');
      img.setAttribute('aria-label', 'Click to enlarge this figure');

      function openLightbox() {
        enlarged.src = img.src;
        enlarged.alt = img.alt || 'Enlarged report figure';
        overlay.classList.add('open');
        closeButton.focus();
      }

      img.addEventListener('click', openLightbox);
      img.addEventListener('keydown', function (event) {
        if (event.key === 'Enter' || event.key === ' ') {
          event.preventDefault();
          openLightbox();
        }
      });
    });

    closeButton.addEventListener('click', closeLightbox);
    overlay.addEventListener('click', function (event) {
      if (event.target === overlay) closeLightbox();
    });
    document.addEventListener('keydown', function (event) {
      if (event.key === 'Escape') closeLightbox();
    });
  }

  if (document.readyState === 'loading') {
    document.addEventListener('DOMContentLoaded', initialiseLightbox);
  } else {
    initialiseLightbox();
  }
})();
</script>
