---
title: "A World of Waves: Why Latitude Explains So Much About the Biggest Waves—but Not Everything"
author: "Julian O'Grady, Tim Leeson and Bryan Hally"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
  word_document:
    toc: true
params:
  aris: [1, 40]
  fit_min_lat: -60
  fit_max_lat: 63
  period_fixed: 120
  a_mode: "equator"
  a_fixed_manual: 4
  equator_band: 5
---

```{r setup, include=FALSE}
# Code is included in HTML but folded closed by default, allowing the
# "Show code" control to reveal it when requested.
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 11,
  fig.height = 6.5,
  fig.align = "center"
)

library(terra)
library(data.table)
library(leaflet)
```

# Introduction

Anyone who follows ocean swell knows that wave heights vary greatly around the world. Latitude—the north–south position on the globe—is a surprisingly powerful influence on this pattern. The energetic storm belts of the mid-latitudes generate more frequent large wind seas and long-travelling swell than most tropical waters. However, latitude cannot explain everything. Coastlines at similar latitudes can have very different wave climates. The greater land area in the Northern Hemisphere breaks up ocean basins and limits fetch and swell pathways, whereas the Southern Ocean provides a more continuous expanse of open water. Wave climates are also shaped by coastal exposure and regional weather systems such as fronts, East Coast Lows and tropical cyclones—also known as hurricanes or typhoons.

This report uses estimates of significant wave height ($H_s$) from a global numerical wave model to explore some key concepts governing how large waves can become. Significant wave height is a statistical description of the sea state, approximately equal to the average height of the highest one-third of waves. It is not the height of the single largest wave, nor does it necessarily represent the breaking-wave face at a surf break.

A probability distribution is fitted at each ocean grid cell to estimate wave heights associated with 1-year and 40-year average recurrence intervals. A deliberately simple latitude-only curve is then fitted to these estimates as a benchmark. The comparison asks: How much of the global pattern can latitude explain, and what additional regional detail does the numerical model provide?

This analysis is not an independent validation against buoy or satellite observations. An example of model validation against wave-buoy observations along [Australia’s east coast is available here](https://ausclimateservice.github.io/buoys2cchaps/buoys2cchaps.html).

The global extreme-wave dataset used here is provided on a relatively coarse 1° grid, which does not resolve detailed nearshore bathymetry or wave transformation close to the coast. Consequently, the mapped values, fitted relationships, statistics and headline findings in this report should be interpreted primarily as representing offshore, deep-water conditions. A follow-up report uses the higher-resolution CCHaPS model and its bathymetry to better represent shallow coastal waters around Australia: [*Shallow Water, Smaller Waves: How Depth Limits Australia’s Biggest Storm Waves*](https://ausclimateservice.github.io/latitude-and-big-waves/CCHaPS_depth_limited_extreme_waves.html).

## A quick guide to the jargon of an ocean-wave climate scientist

| Term | Plain-language meaning |
|:--|:--|
| Significant wave height, $H_s$ | A standard measure of the overall size of a sea state; roughly the average of the highest one-third of waves. |
| ARI (average recurrence interval) | How rarely a level is expected to occur on average. A 40-year ARI is much rarer than a 1-year ARI; it is not a storm lasting 40 years. |
| AEP (annual exceedance probability) | The chance that a level is exceeded at least once in a year under the statistical model. |
| Wave-model-derived estimate | An extreme wave height calculated from a statistical distribution fitted to numerical wave-model output. |
| Latitude-only model | A simple smooth curve that gives every ocean location at the same latitude the same expected extreme $H_s$. |
| Residual | What is left after subtracting the latitude-only value from the full model value. |
| $R^2$ | The share of geographic variation represented by the latitude-only model: 1 is a perfect match and 0 means it explains none of that variation. |

# Aim

This analysis compares wave-model-derived estimates of extreme $H_s$ for **1-year and 40-year ARIs** with a simple cosine curve based only on latitude. In practical terms, it will:

1. calculate the 1-year and 40-year extreme $H_s$ at each ocean grid cell;
2. fit a smooth latitude-only curve to those values;
3. measure how much of the global pattern is represented by latitude alone; and
4. map where the full numerical-model estimate is higher or lower than the latitude-only expectation.

Positive residuals indicate places where the wave-model-derived estimate is larger than the latitude-only expectation. Negative residuals indicate places where it is smaller. Residuals should be interpreted as spatial features represented by the wave-model-derived field, not automatically as physical truth: they may also reflect model resolution, forcing and extreme-value fitting choices.

**Who might find this useful?** This report is intended as an educational explanation for surfers, students, educators and anyone curious about where big waves occur. It provides a simple touchstone for understanding the fundamental controls on the global wave climate---latitude, wind and storm belts, fetch, basin geometry and event rarity---rather than a decision-support or engineering-design product.

# Controls on extreme significant wave height

At the broadest scale, three sets of controls determine the extreme significant wave height likely to occur at a location.

First, consider the storm climate. The Sun heats equatorial regions more strongly than the poles, creating temperature and pressure differences that drive global atmospheric circulation. Earth’s rotation organizes this circulation into prevailing wind belts, while land masses further modify the winds and storm tracks. Even on a water-only Earth with no continents, wave heights would vary with latitude because the tropics, subtropics and mid-latitudes would still experience different winds and storm systems.  

**Read more about global circulation patterns:** the Australian Bureau of Meteorology explains how tropical rising air contributes to the global pressure and wind belts in its [subtropical-ridge explainer](https://www.bom.gov.au/resources/learn-and-explore/climate-knowledge-centre/climate-factors/subtropical-ridge), and describes the storm-bearing Southern Ocean westerlies in its [Southern Annular Mode explainer](https://www.bom.gov.au/climate/sam/).

Second, the real world has **land and ocean basins**. Fetch---the distance over water across which the wind blows in a consistent direction---acts like a runway for wave growth. A long, unobstructed ocean path gives wind more room to build waves, while islands and continents can shelter a coast or block swell arriving from a particular direction. The three classic controls on locally generated wind waves are wind speed, wind duration and fetch ([NOAA National Ocean Service](https://oceanservice.noaa.gov/education/tutorial_currents/03coastal1.html)).

Third, the answer depends on **how rare an event is being considered**. A 40-year ARI significant wave height will generally be larger than a 1-year ARI value because waiting longer gives more opportunity for an unusual storm to occur. This does not mean that the storm lasts for 40 years, or that it arrives regularly every 40 years. ARI describes a long-term average frequency.

Other important controls include swell direction, seasonal and longer-term climate variability, sea ice, and currents that can steepen or focus waves. Near shore, the seabed becomes crucial: depth, refraction, shoaling, bottom friction and breaking transform the offshore sea state before it becomes surf. Because the 1° analysis grid cannot resolve these local processes in detail, this report describes the **offshore, predominantly deep-water extreme wave climate at the model scale**, not the size or quality of waves at a particular surf break.

# Estimating 1-year and 40-year storm waves

This analysis estimates two levels of extreme significant wave height: the 1-year ARI, representing relatively frequent storm waves, and the 40-year ARI, representing much rarer and generally larger storm waves. These are average recurrence intervals rather than schedules: under the statistical convention used here, the chance of exceeding each level in any year is `r formatC(100 * (1 - exp(-1)), digits = 1, format = "f")`% and `r formatC(100 * (1 - exp(-1 / 40)), digits = 2, format = "f")`%, respectively.

The code reads the fitted extreme-value parameters, calculates the two wave-height maps, fits a latitude-only benchmark to each one and masks the displayed rasters to the latitude range used for fitting.

```{r estimate-return-periods, results='hide'}
# Return periods assessed in this report.  Under the Poisson convention,
# AEP = 1 - exp(-1 / ARI), so the 1-year ARI has AEP = 0.6321 rather than 1.
ARIs <- as.numeric(params$aris)
stopifnot(all(is.finite(ARIs)), all(ARIs > 0))

fit_min_lat <- params$fit_min_lat
fit_max_lat <- params$fit_max_lat

# Latitude-only model:
# Hs = a + b * (1 - cos(2*pi*(latitude - phase)/period))
# Keeping the period fixed avoids unstable alternative minima in the nonlinear fit.
period_fixed <- params$period_fixed

# Set the intercept from the equatorial median, or supply a manual value.
a_mode <- params$a_mode
a_fixed_manual <- params$a_fixed_manual
equator_band <- params$equator_band

stopifnot(fit_min_lat < fit_max_lat)
stopifnot(a_mode %in% c("equator", "fixed"))

# Read the GEV parameter field.
ncfile <- "CAWCR_GEVparameters.nc"
data_url <- paste0(
  "https://data-cbr.csiro.au/thredds/fileServer/catch_all/",
  "OA_SLE_archive/COWCLIP/EXTREMES/EVA_R/CSIRO_withMidCentury/",
  "CAWCR_GEVparameters.nc"
)

if (!file.exists(ncfile)) {
  options(timeout = max(600, getOption("timeout")))
  download.file(data_url, ncfile, mode = "wb")
}

r <- rast(ncfile)

# Confirm the layer order against the NetCDF metadata before use.
# The supplied file is assumed to contain: location, scale and shape.
stopifnot(nlyr(r) >= 3)
loc <- r[[1]]
scale <- r[[2]]
shape <- r[[3]]

# Calculate the return levels and fit the latitude-only models.
# Julian's original Hs colour scale.
colz <- rev(c(
  "#000078", "#0000BB", "#0000FF", "#0C3AFC", "#1975FA", "#25B0F7",
  "#32EBF5", "#73EEE6", "#B4F1D7", "#F5F5C8", "#F5F564", "#F5E032",
  "#F6CC00", "#F7A300", "#F97A00", "#FB5100", "#FF0000", "#DA0006",
  "#B5000D", "#910014", "#7B003D", "#660066", "#88008C", "#AB00B2",
  "#CD00D8", "#F000FF", "#CAA6DE", "#DBA7DD", "#EDA9D2"
))
colzRamp <- grDevices::colorRampPalette(rev(colz))
residual_ramp <- grDevices::colorRampPalette(c("#053061", "#2166AC", "#92C5DE", "#F7F7F7", "#F4A582", "#B2182B", "#67001F"))
lat_r <- init(loc, "y")
names(lat_r) <- "latitude"

run_return_period <- function(ARI) {
  AEP <- 1 - exp(-1 / ARI)
  aep_label <- paste0("ARI = ", ARI, " year", if (ARI == 1) "" else "s", "; AEP = ", formatC(AEP, digits = 4, format = "f"))
  y <- -log1p(-AEP)
  Glob_Hs <- ifel(abs(shape) < 1e-8, loc - scale * log(y), loc + scale * (y^(-shape) - 1) / shape)
  names(Glob_Hs) <- paste0("Hs_ARI_", ARI, "yr")
  dt <- as.data.table(as.data.frame(c(Glob_Hs, lat_r), na.rm = TRUE))
  setnames(dt, c("Hs", "latitude"))
  dt <- dt[is.finite(Hs) & is.finite(latitude)]
  lat_q <- dt[, .(p5 = quantile(Hs, .05), p25 = quantile(Hs, .25), p50 = quantile(Hs, .50), p75 = quantile(Hs, .75), p95 = quantile(Hs, .95)), by = latitude]
  setorder(lat_q, latitude)
  fit_dat <- copy(dt[latitude >= fit_min_lat & latitude <= fit_max_lat])
  a_fixed <- if (a_mode == "equator") median(fit_dat[latitude >= -equator_band & latitude <= equator_band, Hs], na.rm = TRUE) else a_fixed_manual
  fit <- nls(Hs ~ a_fixed + b * (1 - cos(2 * pi * (latitude - phase) / period_fixed)), data = fit_dat, start = list(b = 5, phase = 0), algorithm = "port", lower = c(b = 0, phase = -20), upper = c(b = 30, phase = 20), control = nls.control(maxiter = 500, warnOnly = TRUE))
  co <- coef(fit)
  fit_dat[, fitted := predict(fit, newdata = fit_dat)]
  r2_all <- 1 - sum((fit_dat$Hs - fit_dat$fitted)^2) / sum((fit_dat$Hs - mean(fit_dat$Hs))^2)
  lat_q_fit <- copy(lat_q[latitude >= fit_min_lat & latitude <= fit_max_lat])
  lat_q_fit[, fitted := predict(fit, newdata = lat_q_fit)]
  r2_profile <- 1 - sum((lat_q_fit$p50 - lat_q_fit$fitted)^2) / sum((lat_q_fit$p50 - mean(lat_q_fit$p50))^2)
  pred <- data.frame(latitude = seq(fit_min_lat, fit_max_lat, length.out = 500)); pred$fit <- predict(fit, newdata = pred)
  latitude_model <- mask(a_fixed + co[["b"]] * (1 - cos(2 * pi * (lat_r - co[["phase"]]) / period_fixed)), Glob_Hs)
  residual <- Glob_Hs - latitude_model

  # Map versions are restricted to the latitude range used for fitting.
  fit_domain <- lat_r >= fit_min_lat & lat_r <= fit_max_lat
  Hs_map <- ifel(fit_domain, Glob_Hs, NA)
  latitude_model_map <- ifel(fit_domain, latitude_model, NA)
  residual_map <- ifel(fit_domain, residual, NA)

  list(
    ARI = ARI,
    AEP = AEP,
    label = aep_label,
    Hs = Glob_Hs,
    lat_q = lat_q,
    pred = pred,
    a = a_fixed,
    co = co,
    r2_all = r2_all,
    r2_profile = r2_profile,
    latitude_model = latitude_model,
    residual = residual,
    Hs_map = Hs_map,
    latitude_model_map = latitude_model_map,
    residual_map = residual_map
  )
}

results <- lapply(ARIs, run_return_period)

# Named results used throughout the tables, figures and interpretation.
result_1yr  <- results[[match(1, ARIs)]]
result_40yr <- results[[match(40, ARIs)]]
stopifnot(!is.null(result_1yr), !is.null(result_40yr))
```

# Results

## The headline result

The share of the full map-to-map variation represented by latitude alone is measured by the all-cell $R^2$. For example, an $R^2$ of 0.80 means that the latitude-only curve represents 80% of the geographic variance in the full field. The median-profile $R^2$ asks an easier question: how well does the curve follow the typical wave height at each latitude? The difference between the two statistics reflects the east--west and regional structure that a latitude-only curve cannot represent. Both statistics describe the global model grid and therefore primarily characterise offshore, deep-water conditions rather than nearshore wave transformation.

Before looking at the maps, the simple benchmark can be written as

$$
\widehat{H_s}(\phi) = a + b\left[1 - \cos\left(\frac{2\pi(\phi-\phi_0)}{P}\right)\right],
$$

where $\phi$ is latitude, $a$ is the equatorial reference wave height, $b$ is the amplitude controlling how much wave height rises above the tropical baseline, $\phi_0$ is the latitude phase offset that allows a small north--south shift, and $P$ is the fixed period.

This equation is a compact benchmark, not a physical model of every storm. Table 1 gives the numbers that define the two fitted curves. The period is fixed at 120 degrees for both fits so that the comparison is stable and consistent.

```{r fitted-parameter-table}
fit_parameter_table <- do.call(rbind, lapply(results, function(x) {
  data.frame(
    ARI_years = x$ARI,
    AEP = x$AEP,
    equatorial_reference_a_m = x$a,
    amplitude_b_m = unname(x$co[["b"]]),
    phase_degrees = unname(x$co[["phase"]]),
    fixed_period_degrees = period_fixed
  )
}))

knitr::kable(
  fit_parameter_table,
  digits = 3,
  col.names = c(
    "ARI (years)",
    "AEP",
    "Equatorial reference, a (m)",
    "Amplitude, b (m)",
    "Phase (degrees)",
    "Fixed period (degrees)"
  ),
  caption = "Table 1. The fitted parameters that define the latitude-only curves for the 1-year and 40-year ARIs."
)
```

<div class="plain-language-box" role="note">
<div class="plain-language-icon" aria-hidden="true">i</div>
<div class="plain-language-content">
<span class="plain-language-title">In plain language</span>
<p>Across the offshore, predominantly deep-water global analysis grid, latitude alone represents <strong>`r round(100 * result_1yr$r2_all, 1)`%</strong> of the mapped variation in the 1-year extreme wave heights and <strong>`r round(100 * result_40yr$r2_all, 1)`%</strong> in the 40-year extreme wave heights. That is a remarkably large share for one simple geographic variable, but it still leaves important regional differences for the full numerical model to describe.</p>
</div>
</div>

In both cases, more than half of the variance is explained by latitude. This supports the idea that the latitude-dependent distribution of winds and storms is a dominant first-order control. The remaining variance reflects regional influences such as fetch, basin geometry, swell pathways and storm tracks. Latitude therefore describes the broad engine of the global wave climate; it does not tell us exactly how that energy is distributed between individual ocean basins and coastlines.

```{r results-summary}
summary_table <- do.call(rbind, lapply(results, function(x) data.frame(
  ARI_years = x$ARI, AEP = x$AEP, R2_all_grid_cells = x$r2_all,
  variance_explained_all_cells_percent = 100 * x$r2_all,
  R2_median_latitude_profile = x$r2_profile,
  unexplained_all_cells_percent = 100 * (1 - x$r2_all)
)))
knitr::kable(
  summary_table,
  digits = 3,
  col.names = c(
    "ARI (years)",
    "AEP",
    "R²: all grid cells",
    "Variance represented (%)",
    "R²: median latitude profile",
    "Variance not represented (%)"
  ),
  caption = "Table 2. How much of the offshore, predominantly deep-water global wave-height pattern is and is not represented by latitude alone. The raw R² statistics are retained alongside their percentage interpretation."
)
```

```{r return-period-plots, fig.width=12, fig.height=5.6, fig.cap="Figure 1. How wave-model-derived extreme significant wave height changes with latitude for the 1-year ARI (left) and 40-year ARI (right). Both panels use the same vertical scale. At each latitude, the five percentile lines show the range between relatively low- and high-wave regions; the black line is the global median. The purple curve is the latitude-only benchmark, and the grey vertical lines mark the limits used for fitting. These global-model results primarily represent offshore, deep-water conditions. A wide spread between the coloured lines means that places at the same latitude can still have very different wave climates."}
figure1_max_hs <- max(vapply(results, function(x) {
  max(
    c(
      as.matrix(x$lat_q[, .(p5, p25, p50, p75, p95)]),
      x$pred$fit
    ),
    na.rm = TRUE
  )
}, numeric(1)))
figure1_ylim <- c(0, ceiling(figure1_max_hs))

plot_one_result <- function(x) {
  matplot(x$lat_q$latitude, as.matrix(x$lat_q[, .(p5, p25, p50, p75, p95)]), type = "l", lty = c(2, 3, 1, 3, 2), lwd = c(1.5, 1.5, 2.5, 1.5, 1.5), col = c("orange", "red", "black", "dodgerblue", "cyan3"), xlab = "Latitude (degrees)", ylab = "Significant wave height, Hs (m)", ylim = figure1_ylim, main = paste0(x$ARI, "-year ARI extreme Hs\nLatitude represents ", round(100 * x$r2_all, 1), "% of mapped variance"))
  lines(x$pred$latitude, x$pred$fit, col = "purple", lwd = 3); abline(v = c(fit_min_lat, fit_max_lat), col = "grey40")
  legend(
    "topleft",
    legend = c("5th", "25th", "Median", "75th", "95th", "Latitude-only model"),
    lty = c(2, 3, 1, 3, 2, 1),
    lwd = c(1.5, 1.5, 2.5, 1.5, 1.5, 3),
    col = c("orange", "red", "black", "dodgerblue", "cyan3", "purple"),
    ncol = 2,
    bty = "n",
    cex = 0.78
  )
}
old_par <- par(no.readonly = TRUE)
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 4.3, 1.0))
for (x in results) plot_one_result(x)
par(old_par)
```

# Where latitude works—and where it misses

Read Figure 2 from top to bottom. The top row shows the full wave-model-derived estimate: this is the detailed offshore extreme-wave pattern represented on the 1° global analysis grid. The middle row shows what the world would look like if latitude were the only control. The bottom row subtracts the middle from the top, revealing the extra geography represented by the numerical model.

The left column shows the 1-year ARI and the right column shows the rarer 40-year ARI. All six maps use matching colour ranges so they can be compared directly, and they are masked outside the latitude range used to fit the model (`r fit_min_lat`° to `r fit_max_lat`°).

The wave-height colour scale runs from 0 to 18 m in 1 m increments: blues show values below 5 m, yellows and reds show values from 5 to 11 m, and browns, purples and pinks show values above 11 m.

<div class="plain-language-box" role="note">
<div class="plain-language-icon" aria-hidden="true">i</div>
<div class="plain-language-content">
<span class="plain-language-title">In plain language</span>
<p>The regional differences from the latitude-only pattern are generally smaller for the 1-year ARI than for the longer 40-year ARI, whose larger residuals are consistent with a stronger influence from regional extreme-generating systems such as tropical cyclones. The coarse global model may represent broad continental-shelf effects, but it cannot resolve detailed nearshore bathymetry or shallow-water wave transformation. Those processes are examined using a higher-resolution model in the <a href="https://ausclimateservice.github.io/latitude-and-big-waves/CCHaPS_depth_limited_extreme_waves.html"><em>Shallow Water, Smaller Waves</em></a> follow-up report.</p>
</div>
</div>

```{r static-map-comparison, fig.width=12, fig.height=10.5, fig.cap="Figure 2. Where latitude succeeds and where regional geography matters. The 1-year ARI is on the left and the 40-year ARI is on the right. The top row shows the full wave-model-derived estimate, the middle row shows the latitude-only benchmark, and the bottom row shows their difference. Red means the full model is higher than latitude alone would suggest; blue means it is lower. Grey polygons show land. These global-model fields primarily represent offshore, deep-water conditions. The residual patterns highlight the influence of features such as fetch, basin shape, swell pathways and regional storm tracks, as represented by the numerical model."}
old_par <- par(no.readonly = TRUE)
par(
  mfrow = c(3, 2),
  mar = c(1.4, 1.3, 3.2, 3.8),
  oma = c(0.3, 0.3, 0.8, 0.3)
)

# Create a simple land polygon from cells where the ocean wave field is absent.
land_mask_polygon <- as.polygons(
  ifel(is.na(loc), 1, NA),
  aggregate = TRUE,
  values = FALSE,
  na.rm = TRUE
)
land_mask_polygon <- crop(
  land_mask_polygon,
  ext(-180, 180, fit_min_lat, fit_max_lat)
)

add_land_mask <- function() {
  plot(
    land_mask_polygon,
    add = TRUE,
    col = "grey80",
    border = "grey55",
    lwd = 0.3
  )
}

# Top row: full wave-model-derived Hs.
for (x in list(result_1yr, result_40yr)) {
  plot(
    x$Hs_map,
    range = c(0, 18),
    col = colzRamp(18),
    axes = FALSE,
    main = paste0("Wave-model-derived Hs\n", x$label)
  )
  add_land_mask()
}

# Middle row: latitude-only model.
for (x in list(result_1yr, result_40yr)) {
  plot(
    x$latitude_model_map,
    range = c(0, 18),
    col = colzRamp(18),
    axes = FALSE,
    main = paste0("Latitude-only Hs\n", x$label)
  )
  add_land_mask()
}

# Bottom row: numerical-model Hs minus latitude-only Hs.
for (x in list(result_1yr, result_40yr)) {
  plot(
    x$residual_map,
    range = c(-10, 10),
    col = residual_ramp(41),
    axes = FALSE,
    main = paste0("Difference: wave model minus latitude-only\n", x$label)
  )
  add_land_mask()
}

par(old_par)
```

# What surfers and wave-watchers should take from this

- **The maps and statistics primarily represent deep-water, offshore conditions.** The 1° analysis grid does not resolve detailed coastal bathymetry, wave transformation or the final breaking of individual waves experienced at a beach or reef.

- **ARI indicates how rare an event is, not how long you need to wait for it to occur.** A 40-year ARI event is less likely in any given year than a 1-year ARI event. It could occur more than once within 40 years or not occur at all during that period.

- **Latitude sets the broad background.** It places an ocean region within the global wind and storm belts, which is why one simple curve represents more than half of the mapped variance at both return periods.

- **Red and blue regions explain why locations at the same latitude can have different wave heights.** Red indicates that the wave-model-derived estimate is higher than the latitude-only estimate, while blue indicates that it is lower. These differences reflect regional influences such as storm type, fetch, basin geometry, swell pathways, land shelter, sea ice and water depth.

- **Red areas become more prominent in the 40-year difference map.** In tropical-cyclone regions, this is consistent with very extreme tropical-cyclone-generated waves emerging in the rarer tail of the distribution. 

- **Tropical-cyclone wave regions require further modelling.** A latitude-only relationship cannot represent the effects of cyclone intensity, size, forward speed, track or the strongly asymmetric wave field around a cyclone. These regions require dedicated tropical-cyclone wave models, such as the approach described by [O’Grady (2024)](https://bioone.org/journals/journal-of-coastal-research/volume-113/issue-sp1/JCR-SI113-194.1/A-Parametric-Model-of-Tropical-Cyclone-Wave-Fields/10.2112/JCR-SI113-194.1.pdf).

- **Blue areas are often consistent with fetch-limited wave climates.** Enclosed or land-sheltered waters provide less distance for winds to generate waves and can block swell from larger ocean basins. Examples include the North Sea, the Mediterranean Sea and waters off eastern Argentina.

- **Residuals indicate where further investigation is needed.** Red and blue patterns should not automatically be attributed to a single process. They may also reflect model resolution, forcing or statistical uncertainty.

# Explore the maps yourself

Use the radio buttons in the bottom-left corner to switch between the six layers. A useful sequence is to select a wave-model-derived layer, switch to the latitude layer for the same ARI, and then select its residual layer to see where the two differ. Only one data layer is displayed at a time. Pan and zoom to compare ocean basins, remembering that the map primarily represents deep-water, offshore $H_s$, not the final breaking wave at the coast.

*Figure 3. Interactive version of the six fields in Figure 2. Choose a layer in the bottom-left, then pan and zoom to explore the regional pattern. The wave-model-derived and latitude layers use the wave-height colour key in metres. On the residual layers, red means the full model is higher than the latitude-only expectation and blue means it is lower. These global-model fields primarily represent offshore, deep-water conditions.*

```{r interactive-leaflet-map}
if (knitr::is_html_output()) {
  # The slight dateline crop prevents Leaflet projection/wrapping problems.
  leaflet_crop_extent <- ext(rast(xmin = -179.5, xmax = 180))

  prepare_hs_for_leaflet <- function(z) {
    clamp(crop(z, leaflet_crop_extent), lower = 0, upper = 18)
  }

  prepare_residual_for_leaflet <- function(z) {
    clamp(crop(z, leaflet_crop_extent), lower = -10, upper = 10)
  }

  aep_1_label  <- formatC(result_1yr$AEP, digits = 4, format = "f")
  aep_40_label <- formatC(result_40yr$AEP, digits = 4, format = "f")

  interactive_layers <- list(
    prepare_hs_for_leaflet(result_1yr$Hs_map),
    prepare_hs_for_leaflet(result_40yr$Hs_map),
    prepare_hs_for_leaflet(result_1yr$latitude_model_map),
    prepare_hs_for_leaflet(result_40yr$latitude_model_map),
    prepare_residual_for_leaflet(result_1yr$residual_map),
    prepare_residual_for_leaflet(result_40yr$residual_map)
  )

  names(interactive_layers) <- c(
    paste0("Wave-model-derived Hs: ARI 1 (AEP ", aep_1_label, ")"),
    paste0("Wave-model-derived Hs: ARI 40 (AEP ", aep_40_label, ")"),
    paste0("Latitude Hs: ARI 1 (AEP ", aep_1_label, ")"),
    paste0("Latitude Hs: ARI 40 (AEP ", aep_40_label, ")"),
    paste0("Residual: ARI 1 (AEP ", aep_1_label, ")"),
    paste0("Residual: ARI 40 (AEP ", aep_40_label, ")")
  )

  layer_groups <- names(interactive_layers)
  hs_groups <- layer_groups[1:4]
  residual_groups <- layer_groups[5:6]

  hs_pal <- colorNumeric(
    palette = colzRamp(18),
    domain = c(0, 18),
    na.color = "#00000000"
  )

  residual_pal <- colorNumeric(
    palette = residual_ramp(41),
    domain = c(-10, 10),
    na.color = "#00000000"
  )

  interactive_map <- leaflet(
    options = leafletOptions(
      minZoom = 1,
      worldCopyJump = FALSE
    ),
    width = "100%",
    height = 650
  )

  interactive_map <- addProviderTiles(
    interactive_map,
    providers$Esri.WorldGrayCanvas,
    options = providerTileOptions(noWrap = TRUE)
  )

  for (group_name in hs_groups) {
    interactive_map <- addRasterImage(
      interactive_map,
      x = interactive_layers[[group_name]],
      colors = hs_pal,
      opacity = 0.82,
      group = group_name,
      project = TRUE,
      method = "bilinear",
      maxBytes = 8 * 1024 * 1024
    )
  }

  for (group_name in residual_groups) {
    interactive_map <- addRasterImage(
      interactive_map,
      x = interactive_layers[[group_name]],
      colors = residual_pal,
      opacity = 0.82,
      group = group_name,
      project = TRUE,
      method = "bilinear",
      maxBytes = 8 * 1024 * 1024
    )
  }

  # Add only two colour keys: one shared by all Hs layers and one shared by
  # both residual layers.
  interactive_map <- addLegend(
    interactive_map,
    pal = hs_pal,
    values = c(0, 18),
    title = "Hs (m)",
    opacity = 1,
    position = "bottomright"
  )

  interactive_map <- addLegend(
    interactive_map,
    pal = residual_pal,
    values = c(-10, 10),
    title = "Residual Hs (m)",
    opacity = 1,
    position = "bottomright"
  )

  # baseGroups produces radio buttons, so only one raster is selected at a time.
  interactive_map <- addLayersControl(
    interactive_map,
    baseGroups = layer_groups,
    options = layersControlOptions(
      collapsed = FALSE,
      position = "bottomleft"
    )
  )

  # Begin with the 1-year wave-model-derived field displayed.
  for (group_name in layer_groups[-1]) {
    interactive_map <- hideGroup(interactive_map, group_name)
  }

  interactive_map <- addScaleBar(
    interactive_map,
    position = "topright",
    options = scaleBarOptions(metric = TRUE, imperial = FALSE)
  )

  interactive_map <- fitBounds(
    interactive_map,
    lng1 = -179.5,
    lat1 = fit_min_lat,
    lng2 = 180,
    lat2 = fit_max_lat
  )

  interactive_map
} else {
  cat("The interactive map is available in the HTML version of this report.")
}
```

## Spin the same layers on a 3D globe

The globe uses the same six masked rasters and the same colour scales as the flat map above. Choose the wave-model-derived field, latitude-only model or residual, then switch between the 1-year and 40-year ARIs. Drag to rotate the globe, use the mouse wheel or trackpad to zoom, and pause or restart the automatic spin with the button.

*Figure 4. Interactive 3D view of the same wave-height fields. The coloured ocean surface is restricted to the latitude range used for fitting and primarily represents offshore, deep-water conditions. Grey areas are land or locations outside that fitted range. The legend changes automatically between significant wave height and residual wave height.*

```{r interactive-threejs-globe, results='asis'}
if (knitr::is_html_output()) {
  stopifnot(exists("interactive_map"), length(layer_groups) == 6)

  globe_id <- paste0("wave-globe-", sprintf("%08x", sample.int(.Machine$integer.max, 1)))

  globe_config <- list(
    rootId = globe_id,
    groups = stats::setNames(
      as.list(layer_groups),
      c("gev-1", "gev-40", "latitude-1", "latitude-40", "residual-1", "residual-40")
    ),
    palettes = list(
      hs = unname(colzRamp(18)),
      residual = unname(residual_ramp(41))
    ),
    fitLatitude = c(fit_min_lat, fit_max_lat)
  )

  globe_css <- paste0(
    "#", globe_id, " { position: relative; width: 100%; min-height: 650px; overflow: hidden; background: #eeeeee; border: 1px solid #c7c7c7; border-radius: 6px; }\n",
    "#", globe_id, " .wave-globe-canvas { display: block; width: 100%; height: 650px; cursor: grab; touch-action: none; }\n",
    "#", globe_id, " .wave-globe-canvas:active { cursor: grabbing; }\n",
    "#", globe_id, " .wave-globe-controls { position: absolute; z-index: 2; top: 12px; left: 12px; max-width: calc(100% - 24px); padding: 10px; color: #263238; background: rgba(255,255,255,0.92); border: 1px solid rgba(70,70,70,0.28); border-radius: 5px; box-shadow: 0 1px 4px rgba(0,0,0,0.15); }\n",
    "#", globe_id, " .wave-globe-control-row { display: flex; flex-wrap: wrap; align-items: center; gap: 6px; margin-bottom: 7px; }\n",
    "#", globe_id, " .wave-globe-control-row:last-child { margin-bottom: 0; }\n",
    "#", globe_id, " .wave-globe-label { min-width: 48px; font-size: 13px; font-weight: 700; }\n",
    "#", globe_id, " button { padding: 5px 9px; color: #263238; background: #f7f7f7; border: 1px solid #8a9a9e; border-radius: 4px; font: inherit; font-size: 13px; cursor: pointer; }\n",
    "#", globe_id, " button:hover { background: #e7f3f7; }\n",
    "#", globe_id, " button.is-active { color: #ffffff; background: #087da8; border-color: #087da8; }\n",
    "#", globe_id, " button:focus-visible { outline: 3px solid #45a9d1; outline-offset: 2px; }\n",
    "#", globe_id, " .wave-globe-status { max-width: 410px; font-size: 12px; }\n",
    "#", globe_id, " .wave-globe-legend { position: absolute; z-index: 2; right: 12px; bottom: 12px; width: min(280px, calc(100% - 24px)); padding: 9px 10px; color: #263238; background: rgba(255,255,255,0.92); border: 1px solid rgba(70,70,70,0.28); border-radius: 5px; box-shadow: 0 1px 4px rgba(0,0,0,0.15); }\n",
    "#", globe_id, " .wave-globe-legend-title { margin-bottom: 5px; font-size: 13px; font-weight: 700; }\n",
    "#", globe_id, " .wave-globe-legend-bar { height: 14px; border: 1px solid rgba(40,40,40,0.45); }\n",
    "#", globe_id, " .wave-globe-legend-ticks { display: flex; justify-content: space-between; margin-top: 3px; font-size: 12px; }\n",
    "@media (max-width: 600px) { #", globe_id, " { min-height: 590px; } #", globe_id, " .wave-globe-canvas { height: 590px; } #", globe_id, " .wave-globe-controls { right: 12px; } }\n"
  )

  globe_markup <- htmltools::tagList(
    htmltools::tags$style(htmltools::HTML(globe_css)),
    htmltools::tags$div(
      id = globe_id,
      class = "wave-globe-shell",
      markdown = "0",
      htmltools::tags$div(
        class = "wave-globe-controls",
        htmltools::tags$div(
          class = "wave-globe-control-row",
          htmltools::tags$span(class = "wave-globe-label", "Layer"),
          htmltools::tags$button(type = "button", class = "wave-model is-active", `data-model` = "gev", `aria-pressed` = "true", "Wave-model-derived"),
          htmltools::tags$button(type = "button", class = "wave-model", `data-model` = "latitude", `aria-pressed` = "false", "Latitude-only"),
          htmltools::tags$button(type = "button", class = "wave-model", `data-model` = "residual", `aria-pressed` = "false", "Residual")
        ),
        htmltools::tags$div(
          class = "wave-globe-control-row",
          htmltools::tags$span(class = "wave-globe-label", "ARI"),
          htmltools::tags$button(type = "button", class = "wave-ari is-active", `data-ari` = "1", `aria-pressed` = "true", "1 year"),
          htmltools::tags$button(type = "button", class = "wave-ari", `data-ari` = "40", `aria-pressed` = "false", "40 years"),
          htmltools::tags$button(type = "button", class = "wave-spin", `aria-pressed` = "true", "Pause spin")
        ),
        htmltools::tags$div(class = "wave-globe-status", `aria-live` = "polite", "Loading 3D globe...")
      ),
      htmltools::tags$canvas(
        class = "wave-globe-canvas",
        role = "img",
        `aria-label` = "Interactive spinning globe showing modelled extreme significant wave height"
      ),
      htmltools::tags$div(
        class = "wave-globe-legend",
        htmltools::tags$div(class = "wave-globe-legend-title", "Significant wave height, Hs (m)"),
        htmltools::tags$div(class = "wave-globe-legend-bar"),
        htmltools::tags$div(
          class = "wave-globe-legend-ticks",
          htmltools::tags$span("0"),
          htmltools::tags$span("5"),
          htmltools::tags$span("11"),
          htmltools::tags$span("18")
        )
      )
    )
  )

  globe_javascript <- paste0(
    "const waveGlobeConfig = ",
    jsonlite::toJSON(globe_config, auto_unbox = TRUE, null = "null"),
    ";\n",
    r"---(
const root = document.getElementById(waveGlobeConfig.rootId);

if (root) {
  const canvas = root.querySelector(".wave-globe-canvas");
  const status = root.querySelector(".wave-globe-status");
  const legendTitle = root.querySelector(".wave-globe-legend-title");
  const legendBar = root.querySelector(".wave-globe-legend-bar");
  const legendTicks = root.querySelector(".wave-globe-legend-ticks");
  const spinButton = root.querySelector(".wave-spin");

  const state = {
    model: "gev",
    ari: "1",
    spinning: !window.matchMedia("(prefers-reduced-motion: reduce)").matches
  };

  if (!state.spinning) {
    spinButton.textContent = "Start spin";
    spinButton.setAttribute("aria-pressed", "false");
  }

  function findLeafletRasters() {
    const expectedGroups = Object.values(waveGlobeConfig.groups);
    const widgetScripts = Array.from(document.querySelectorAll("script[data-for]"));

    for (const script of widgetScripts) {
      let payload;
      try {
        payload = JSON.parse(script.textContent);
      } catch (error) {
        continue;
      }

      const calls = payload && payload.x && Array.isArray(payload.x.calls)
        ? payload.x.calls
        : [];
      const rasterCalls = calls.filter(call => call.method === "addRasterImage");
      const byGroup = new Map();

      rasterCalls.forEach(call => {
        if (call.args && call.args.length >= 4) {
          byGroup.set(call.args[3], {
            uri: call.args[0],
            bounds: call.args[1]
          });
        }
      });

      if (expectedGroups.every(group => byGroup.has(group))) {
        return byGroup;
      }
    }

    throw new Error("The raster layers from the Leaflet map could not be found.");
  }

  function paletteGradient(colours) {
    return "linear-gradient(to right, " + colours.join(", ") + ")";
  }

  function setLegend() {
    const residual = state.model === "residual";
    legendTitle.textContent = residual
      ? "Residual wave height (m)"
      : "Significant wave height, Hs (m)";
    legendBar.style.background = paletteGradient(
      residual ? waveGlobeConfig.palettes.residual : waveGlobeConfig.palettes.hs
    );
    legendTicks.innerHTML = residual
      ? "<span>-10</span><span>-5</span><span>0</span><span>5</span><span>10</span>"
      : "<span>0</span><span>5</span><span>11</span><span>18</span>";
  }

  function setPressed(buttons, activeButton) {
    buttons.forEach(button => {
      const active = button === activeButton;
      button.classList.toggle("is-active", active);
      button.setAttribute("aria-pressed", active ? "true" : "false");
    });
  }

  try {
    const rastersByGroup = findLeafletRasters();
    const THREE = await import("https://esm.sh/three@0.169.0");
    const controlsModule = await import("https://esm.sh/three@0.169.0/examples/jsm/controls/OrbitControls.js");
    const OrbitControls = controlsModule.OrbitControls;

    const renderer = new THREE.WebGLRenderer({ canvas, antialias: true, alpha: false });
    renderer.setPixelRatio(Math.min(window.devicePixelRatio || 1, 2));
    renderer.outputColorSpace = THREE.SRGBColorSpace;
    renderer.setClearColor(0xeeeeee, 1);

    const scene = new THREE.Scene();
    const camera = new THREE.PerspectiveCamera(35, 1, 0.1, 100);
    camera.position.set(0, 0.15, 3.35);

    const controls = new OrbitControls(camera, canvas);
    controls.enableDamping = true;
    controls.dampingFactor = 0.06;
    controls.enablePan = false;
    controls.minDistance = 2.15;
    controls.maxDistance = 6.0;

    const globeGroup = new THREE.Group();
    globeGroup.rotation.y = THREE.MathUtils.degToRad(125);
    scene.add(globeGroup);

    const baseSphere = new THREE.Mesh(
      new THREE.SphereGeometry(1, 128, 64),
      new THREE.MeshStandardMaterial({
        color: 0xd2d2d2,
        roughness: 0.92,
        metalness: 0
      })
    );
    globeGroup.add(baseSphere);

    const overlayMaterial = new THREE.MeshBasicMaterial({
      transparent: true,
      opacity: 0.96,
      depthWrite: false
    });
    const overlaySphere = new THREE.Mesh(
      new THREE.SphereGeometry(1.006, 128, 64),
      overlayMaterial
    );
    globeGroup.add(overlaySphere);

    scene.add(new THREE.HemisphereLight(0xffffff, 0x777777, 1.55));
    const keyLight = new THREE.DirectionalLight(0xffffff, 1.05);
    keyLight.position.set(3, 2, 4);
    scene.add(keyLight);

    function latitudeToPoint(latitude, longitude, radius) {
      const phi = THREE.MathUtils.degToRad(90 - latitude);
      const theta = THREE.MathUtils.degToRad(longitude + 180);
      return new THREE.Vector3(
        -radius * Math.sin(phi) * Math.cos(theta),
        radius * Math.cos(phi),
        radius * Math.sin(phi) * Math.sin(theta)
      );
    }

    const gridMaterial = new THREE.LineBasicMaterial({
      color: 0x8f8f8f,
      transparent: true,
      opacity: 0.34
    });

    for (let latitude = -60; latitude <= 60; latitude += 30) {
      const points = [];
      for (let longitude = -180; longitude <= 180; longitude += 2) {
        points.push(latitudeToPoint(latitude, longitude, 1.012));
      }
      globeGroup.add(new THREE.Line(new THREE.BufferGeometry().setFromPoints(points), gridMaterial));
    }

    for (let longitude = -150; longitude < 180; longitude += 30) {
      const points = [];
      for (let latitude = -90; latitude <= 90; latitude += 2) {
        points.push(latitudeToPoint(latitude, longitude, 1.012));
      }
      globeGroup.add(new THREE.Line(new THREE.BufferGeometry().setFromPoints(points), gridMaterial));
    }

    const textureCache = new Map();

    function mercatorY(latitude) {
      const limited = Math.max(-85.05112878, Math.min(85.05112878, latitude));
      const radians = THREE.MathUtils.degToRad(limited);
      return Math.log(Math.tan(Math.PI / 4 + radians / 2));
    }

    async function makeEquirectangularTexture(raster) {
      if (textureCache.has(raster.uri)) return textureCache.get(raster.uri);

      const image = new Image();
      image.src = raster.uri;
      await image.decode();

      const textureCanvas = document.createElement("canvas");
      textureCanvas.width = 1440;
      textureCanvas.height = 720;
      const context = textureCanvas.getContext("2d");
      context.clearRect(0, 0, textureCanvas.width, textureCanvas.height);
      context.imageSmoothingEnabled = true;
      context.imageSmoothingQuality = "high";

      const flatBounds = raster.bounds.flat ? raster.bounds.flat() : [].concat(...raster.bounds);
      const north = Number(flatBounds[0]);
      const south = Number(flatBounds[2]);
      const mercatorNorth = mercatorY(north);
      const mercatorSouth = mercatorY(south);

      for (let outputY = 0; outputY < textureCanvas.height; outputY += 1) {
        const latitude = 90 - 180 * ((outputY + 0.5) / textureCanvas.height);
        if (latitude > north || latitude < south) continue;

        const sourceFraction = (mercatorNorth - mercatorY(latitude)) /
          (mercatorNorth - mercatorSouth);
        const sourceY = Math.max(
          0,
          Math.min(image.height - 1, Math.floor(sourceFraction * image.height))
        );
        context.drawImage(
          image,
          0,
          sourceY,
          image.width,
          1,
          0,
          outputY,
          textureCanvas.width,
          1
        );
      }

      const texture = new THREE.CanvasTexture(textureCanvas);
      texture.colorSpace = THREE.SRGBColorSpace;
      texture.anisotropy = Math.min(8, renderer.capabilities.getMaxAnisotropy());
      texture.needsUpdate = true;
      textureCache.set(raster.uri, texture);
      return texture;
    }

    function currentConfigKey() {
      return state.model + "-" + state.ari;
    }

    async function updateLayer() {
      const configKey = currentConfigKey();
      const groupName = waveGlobeConfig.groups[configKey];
      const raster = rastersByGroup.get(groupName);

      status.textContent = "Loading " + groupName + "...";
      overlayMaterial.map = await makeEquirectangularTexture(raster);
      overlayMaterial.needsUpdate = true;
      setLegend();
      status.textContent = groupName;
    }

    root.querySelectorAll(".wave-model").forEach(button => {
      button.addEventListener("click", async () => {
        state.model = button.dataset.model;
        setPressed(root.querySelectorAll(".wave-model"), button);
        await updateLayer();
      });
    });

    root.querySelectorAll(".wave-ari").forEach(button => {
      button.addEventListener("click", async () => {
        state.ari = button.dataset.ari;
        setPressed(root.querySelectorAll(".wave-ari"), button);
        await updateLayer();
      });
    });

    spinButton.addEventListener("click", () => {
      state.spinning = !state.spinning;
      spinButton.textContent = state.spinning ? "Pause spin" : "Start spin";
      spinButton.setAttribute("aria-pressed", state.spinning ? "true" : "false");
    });

    function resizeRenderer() {
      const width = Math.max(1, canvas.clientWidth);
      const height = Math.max(1, canvas.clientHeight);
      const pixelRatio = renderer.getPixelRatio();
      const requiredWidth = Math.floor(width * pixelRatio);
      const requiredHeight = Math.floor(height * pixelRatio);

      if (canvas.width !== requiredWidth || canvas.height !== requiredHeight) {
        renderer.setSize(width, height, false);
        camera.aspect = width / height;
        camera.updateProjectionMatrix();
      }
    }

    const resizeObserver = new ResizeObserver(resizeRenderer);
    resizeObserver.observe(root);

    await updateLayer();

    let previousTime = performance.now();
    function animate(now) {
      const elapsedSeconds = Math.min(0.05, (now - previousTime) / 1000);
      previousTime = now;
      resizeRenderer();
      controls.update();
      if (state.spinning) globeGroup.rotation.y += elapsedSeconds * 0.10;
      renderer.render(scene, camera);
      requestAnimationFrame(animate);
    }
    requestAnimationFrame(animate);
  } catch (error) {
    status.textContent = "The 3D globe could not be loaded: " + error.message;
    console.error(error);
  }
}
)---"
  )

  full_globe_markup <- htmltools::tagList(
    globe_markup,
    htmltools::tags$script(
      type = "module",
      htmltools::HTML(globe_javascript)
    )
  )

  # htmltools indents nested tags when converting them to text. Pandoc can
  # interpret those indented child tags as Markdown code blocks, which prints
  # the controls as literal HTML. Collapse whitespace between tags so the
  # complete widget is passed through as raw HTML.
  globe_html <- paste(as.character(full_globe_markup), collapse = "")
  globe_html <- gsub(">\\s+<", "><", globe_html, perl = TRUE)
  cat(globe_html, "\n")
} else {
  cat("The interactive 3D globe is available in the HTML version of this report.")
}
```

```{=html}
<style>
  .plain-language-box {
    display: flex;
    align-items: flex-start;
    gap: 14px;
    margin: 1.25em 0;
    padding: 16px 18px;
    color: #173541;
    background: #eaf7fb;
    border: 1px solid #b8dfe8;
    border-left: 6px solid #00a7c4;
    border-radius: 7px;
  }

  .plain-language-icon {
    display: flex;
    flex: 0 0 30px;
    align-items: center;
    justify-content: center;
    width: 30px;
    height: 30px;
    margin-top: 1px;
    color: white;
    background: #087da8;
    border-radius: 50%;
    font-size: 20px;
    font-weight: 700;
    line-height: 1;
  }

  .plain-language-content {
    flex: 1 1 auto;
    min-width: 0;
    font-size: 1.14em;
    line-height: 1.5;
  }

  .plain-language-content p {
    margin: 0;
  }

  .plain-language-title {
    display: block;
    margin-bottom: 0.2em;
    font-size: 1.04em;
    font-weight: 700;
  }

  .figure img {
    cursor: zoom-in;
  }

  .figure img:focus {
    outline: 3px solid #3498db;
    outline-offset: 4px;
  }

  #plot-lightbox {
    display: none;
    position: fixed;
    inset: 0;
    z-index: 10000;
    align-items: center;
    justify-content: center;
    padding: 54px 28px 28px;
    background: rgba(0, 0, 0, 0.9);
  }

  #plot-lightbox.is-open {
    display: flex;
  }

  #plot-lightbox-image {
    display: block;
    max-width: 96vw;
    max-height: 90vh;
    width: auto;
    height: auto;
    object-fit: contain;
    background: white;
  }

  #plot-lightbox-close {
    position: absolute;
    top: 10px;
    right: 18px;
    border: 0;
    padding: 2px 12px 8px;
    color: white;
    background: transparent;
    font-size: 42px;
    line-height: 1;
    cursor: pointer;
  }

  #plot-lightbox-close:focus {
    outline: 3px solid white;
    outline-offset: 2px;
  }
</style>

<div id="plot-lightbox" role="dialog" aria-modal="true" aria-label="Enlarged figure" aria-hidden="true">
  <button id="plot-lightbox-close" type="button" aria-label="Close enlarged figure">&times;</button>
  <img id="plot-lightbox-image" alt="">
</div>

<script>
(function () {
  function initialisePlotLightbox() {
    var lightbox = document.getElementById("plot-lightbox");
    var enlargedImage = document.getElementById("plot-lightbox-image");
    var closeButton = document.getElementById("plot-lightbox-close");
    var previousFocus = null;

    if (!lightbox || !enlargedImage || !closeButton) return;

    function openLightbox(sourceImage) {
      previousFocus = sourceImage;
      enlargedImage.src = sourceImage.src;
      enlargedImage.alt = sourceImage.alt || "Enlarged figure";
      lightbox.classList.add("is-open");
      lightbox.setAttribute("aria-hidden", "false");
      document.body.style.overflow = "hidden";
      closeButton.focus();
    }

    function closeLightbox() {
      lightbox.classList.remove("is-open");
      lightbox.setAttribute("aria-hidden", "true");
      enlargedImage.removeAttribute("src");
      document.body.style.overflow = "";
      if (previousFocus) previousFocus.focus();
    }

    document.querySelectorAll(".figure img").forEach(function (image) {
      image.setAttribute("tabindex", "0");
      image.setAttribute("role", "button");
      image.setAttribute("aria-label", "Enlarge figure: " + (image.alt || "plot"));

      image.addEventListener("click", function () {
        openLightbox(image);
      });

      image.addEventListener("keydown", function (event) {
        if (event.key === "Enter" || event.key === " ") {
          event.preventDefault();
          openLightbox(image);
        }
      });
    });

    closeButton.addEventListener("click", closeLightbox);

    lightbox.addEventListener("click", function (event) {
      if (event.target === lightbox) closeLightbox();
    });

    document.addEventListener("keydown", function (event) {
      if (event.key === "Escape" && lightbox.classList.contains("is-open")) {
        closeLightbox();
      }
    });
  }

  if (document.readyState === "loading") {
    document.addEventListener("DOMContentLoaded", initialisePlotLightbox);
  } else {
    initialisePlotLightbox();
  }
})();
</script>
```

# Looking ahead: a changing global circulation

Global atmospheric circulation patterns are changing as the climate warms; for example, the joint CSIRO and Bureau of Meteorology [*State of the Climate 2024*](https://www.csiro.au/-/media/Environment/SOTC-2024/State-of-the-Climate-2024-Accessible-version.html) reports climate-related changes in atmospheric pressure and high-pressure systems across southern Australia. Because wave climates respond to the position and strength of wind belts, storm tracks and tropical-cyclone activity, these changes could alter both the fitted latitude relationship and the regional residual patterns. This does not invalidate the historical analysis, but its results should not be assumed to remain fixed: future applications should rerun the method using updated observations or future-climate wave-model projections.

# Where the wave numbers come from

The analysis uses a global set of extreme-value parameters derived from the CAWCR numerical wave model. It is not a live forecast, a direct buoy record or a surf-spot model. The extreme-value dataset is gridded at approximately 1° resolution and is therefore suited to broad offshore and deep-water patterns rather than detailed coastal or shallow-water conditions. The code downloads `CAWCR_GEVparameters.nc` from the CSIRO THREDDS server:

<https://data-cbr.csiro.au/thredds/fileServer/catch_all/OA_SLE_archive/COWCLIP/EXTREMES/EVA_R/CSIRO_withMidCentury/CAWCR_GEVparameters.nc>

Direct input-file citation: CSIRO. (n.d.). *CAWCR_GEVparameters.nc* [NetCDF data file]. COWCLIP / EXTREMES / EVA_R / CSIRO_withMidCentury. Retrieved 17 August 2026 from the URL above.

The broader processed global extreme-wave data collection and methodology relevant to this input should also be cited as follows:

O'Grady, J. G., Hemer, M. A., McInnes, K. L., Trenham, C. E., Stephenson, A. G., & Seers, B. (2021). *Projected incremental changes to extreme wind-driven wave heights for the twenty-first century* (Version 1) [Data collection]. CSIRO. <https://doi.org/10.25919/82cw-8w26>

O'Grady, J. G., Hemer, M. A., McInnes, K. L., Trenham, C. E., Stephenson, A. G., & Seers, B. (2021). Projected incremental changes to extreme wind-driven wave heights for the twenty-first century. *Scientific Reports, 11*, 8826. <https://doi.org/10.1038/s41598-021-87358-w>

The CAWCR global wave model is documented separately by Smith et al. (2020), who describe the global hindcast, its forcing and spatial resolution: <https://doi.org/10.1002/gdj3.104>.

```{r save-png, include=FALSE}
for (x in results) {
  png_file <- paste0("Hs_ARI_", x$ARI, "yr_percentiles_by_latitude_all_cells_cosine_fit.png")
  grDevices::png(png_file, width = 1500, height = 950, res = 130)
  plot_one_result(x)
  grDevices::dev.off()
  cat("Saved figure: ", png_file, "\n", sep = "")
}
```

# Limits and reproducibility notes

- The results, fitted relationships and $R^2$ statistics primarily describe offshore, deep-water significant wave height at the approximately 1° analysis-grid scale. They do not represent breaking-wave face height, surf quality or detailed wave transformation at an individual beach or reef.
- The raster contains fitted GEV parameters rather than direct observations. `Glob_Hs` is therefore a **wave-model-derived estimate** calculated from those parameters.
- The higher-resolution CCHaPS analysis in [*Shallow Water, Smaller Waves*](https://ausclimateservice.github.io/latitude-and-big-waves/CCHaPS_depth_limited_extreme_waves.html) provides a more suitable representation of Australian coastal waters and shallower depths.

Future analysis could test whether the residual patterns are associated with tropical-cyclone track exposure, distance from the coast or bathymetric depth.
